---
title: "Comparison of variance decomposition using GFA, iCLuster and MOFA"
author: "Britta Velten"
date: '`r format(Sys.time(), "%d %B, %Y")`'
output:
  BiocStyle::html_document:
    toc: true
---

```{r, warning=F, message=F}
setwd("/Users/bvelten/Documents/MOFA/CLL_MOFA_data/Analysis/code4manuscript/iCluster/")
library(iClusterPlus)
library(MOFAtools)
library(ggplot2)
library(cowplot)
library(pheatmap)
library(gridExtra)
```

# I/O Options
The input folders point to directories with the simulated data and the fitted MOFA, GFA and iCluster models.
```{r}
in.folder <- "/Users/bvelten/Documents/MOFA/CLL_MOFA_data/Analysis/code4manuscript/iCluster/Simulations/data/"
iClfit.folder <- "/Users/bvelten/Documents/MOFA/CLL_MOFA_data/Analysis/code4manuscript/iCluster/Simulations/iCluster_results/"
MOFAfit.folder <- "/Users/bvelten/Documents/MOFA/CLL_MOFA_data/Analysis/code4manuscript/iCluster/Simulations/MOFA_results/"
GFAfitfolder <- "/Users/bvelten/Documents/MOFA/CLL_MOFA_data/Analysis/code4manuscript/iCluster/Simulations/GFA_results/"

plotdir <- "figures/r2_comparison/"
if(!dir.exists(plotdir)) dir.create(plotdir)
knitr::opts_chunk$set(fig.path = plotdir, dev=c("png", "pdf"))
```

# Load simulated data
```{r}
files <- list.files(in.folder, pattern=".txt$")

# data
simData <- lapply(files, function(fnm){
  dt <- read.table(file.path(in.folder,fnm))
  as.matrix(dt)
})


# view index
mvec <- sapply(files, function(fnm){
	  split <- strsplit(fnm,"_")[[1]]
  	substr(split[[2]],1,nchar(split[[2]])-4)
	})

# number of facotrs
kvec <- sapply(files, function(fnm){
	split <- strsplit(fnm,"_")[[1]]
	split[1]
	})

# set names
names(simData) <- paste("k",kvec,"_","m",mvec, sep="")
#center
simData[grepl("m[0-2]", names(simData))] <- lapply(simData[grepl("m[0-2]", names(simData))], function(dd) scale(dd, center=TRUE, scale=FALSE))

```

# Only consider k=10,15
```{r}
simData <- simData[grepl("k10|k15", names(simData))]
kvec <- kvec[grepl("10_|15_", names(kvec))]
```

# Run iCluster (slow, done on server)
```{r, eval=F}
# fit iCluster model
for(k in unique(kvec)){
tuned.out <- tune.iClusterPlus(cpus =12, dt1=simData[[paste("k",k,"_","m",0, sep="")]],
                               dt2=simData[[paste("k",k,"_","m",1, sep="")]],
                               dt3=simData[[paste("k",k,"_","m",2, sep="")]],
                               type=c("gaussian","gaussian","gaussian"),
                               K=as.numeric(k), n.lambda=185, scale.lambda=c(1,1,1),maxiter=20)
#save result
save(tuned.out, file=paste("iclustertuned.out",k,".RData", sep=""))
}
```

# Load fitted models (iCluster)
```{r}
iCluster.out <- lapply(unique(kvec), function(k) {
  load(file.path(iClfit.folder,paste("iclustertuned.out",k,".RData", sep="")))
  tuned.out
  })
names(iCluster.out) <- unique(kvec)
```

# Load fitted models (MOFA)
```{r}
MOFA.out <- lapply(unique(kvec), function(k) loadModel(file.path(MOFAfit.folder,paste("0_",k,".hdf5", sep=""))))
names(MOFA.out) <- unique(kvec)
```

# Load fitted models (GFA)
```{r}
GFA.out <- lapply(unique(kvec), function(k) readRDS(file.path(GFAfitfolder,paste("gfa_ktrue",k,"_kstart25.rds", sep=""))))
names(GFA.out) <- unique(kvec)
```

# Load fitted models (GFA_fixed)
```{r}
GFA_fixed.out <- lapply(unique(kvec), function(k) readRDS(file.path(GFAfitfolder,paste("gfa_ktrue",k,"_kstart",k,".rds", sep=""))))
names(GFA_fixed.out) <- unique(kvec)
```

# True alpha
```{r}
alpha <- simData[grepl("alpha", names(simData))]
names(alpha) <- sub("k","",sapply(strsplit(names(alpha), "_"), "[[",1))
alpha <- alpha[unique(kvec)]
```

# Helper functions
Function to calculate variance explained per view for iCluster 'calculateVarianceExplained4iClust'
```{r, echo=F}
calculateVarianceExplained4iClust <- function(iCl_fit, k=NULL, data, plotit=T, perFeature=F, 
                                       orderFactorsbyR2=F, showtotalR2=T, showVarComp=F) {
  
  # if only fit for one k transform to list
  if(!is.null(names(iCl_fit)) & all(names(iCl_fit)==c("fit","lambda"))) iCl_fit <- list(iCl_fit)
  
  kvec = as.character(sapply(iCl_fit, function(out) ncol(out$fit[[1]]$meanZ)))
  names(iCl_fit) <- kvec
  
  if(is.null(k)) k <- kvec[1]
  print(paste("Calculating for k=", k))
  stopifnot(k %in% kvec)
  
  #choose best model based on BIC
  bic <-  getBIC(iCl_fit)
  bestFitIdx <- apply(bic,2,which.min)
  names(bestFitIdx) <- kvec
  bestFit <- lapply(kvec, function(i) iCl_fit[[i]]$fit[[bestFitIdx[i]]])
  names(bestFit) <-kvec

  #choose model with specified k
  iClmodel <- bestFit[[k]]
  weightsiClust <- iClmodel$beta
  Y <- data[grepl(paste("k",k,"_m[0-2]", sep=""),names(data))]
  ZiClust <- iClmodel$meanZ
  # corrplot::corrplot(cor(ZiClust), title = "iCluster factors correlation")
  intercept <- iClmodel$alpha
  names(intercept) <- names(Y)
  SW <- lapply(seq_along(weightsiClust), function(d) cbind(intercept=intercept[[d]], weightsiClust[[d]])) 
  names(SW) <- names(Y)
  Z <- cbind(intercept=1,ZiClust)
  colnames(Z) <- c("intercept", paste(1:ncol(ZiClust), "iClust", sep="_"))
  for (i in seq_along(SW)) colnames(SW[[i]]) <- c("intercept", paste(1:ncol(ZiClust), "iClust", sep="_"))
  factors <- colnames(Z)

  factors <- factors[factors!="intercept"]
  
  K <- length(factors)
  views <-names(SW)
  #iCluster always fits an intercept, but as the data is centered this is not necessar and leads to negative R@ as iCluster does not learn the intercept perfectly, same when using uncentered data: Hence exclude the learnt intercept
  include_intercept <- FALSE
  
  ### copied from claculate variance explained mofa function
  FeatureMean <- lapply(views, function(m) {
        apply(Y[[m]], 2, mean, na.rm = T)
    })
    names(FeatureMean) <- views
    resNullModel <- lapply(views, function(m) sweep(Y[[m]], 2, 
        FeatureMean[[m]], "-"))
    names(resNullModel) <- views
    Z[is.na(Z)] <- 0
    Ypred_m <- lapply(views, function(m) Z %*% t(SW[[m]]))
    names(Ypred_m) <- views
    Ypred_mk <- lapply(views, function(m) {
        ltmp <- lapply(factors, function(k) Z[, k] %*% t(SW[[m]][, 
            k]))
        names(ltmp) <- factors
        ltmp
    })
    names(Ypred_mk) <- views
    if (include_intercept) {
        Y <- lapply(views, function(m) sweep(Y[[m]], 2, intercept[[m]], 
            "-"))
        names(Y) <- views
    }
    fvar_m <- sapply(views, function(m) 1 - sum((Y[[m]] - Ypred_m[[m]])^2, 
        na.rm = T)/sum(resNullModel[[m]]^2, na.rm = T))
    fvar_mk <- matrix(sapply(views, function(m) sapply(factors, 
        function(k) 1 - sum((Y[[m]] - Ypred_mk[[m]][[k]])^2, 
            na.rm = T)/sum(resNullModel[[m]]^2, na.rm = T))), 
        ncol = length(views), nrow = length(factors))
    names(fvar_m) <- views
    colnames(fvar_mk) <- views
    rownames(fvar_mk) <- factors
    R2_list <- list(R2Total = fvar_m, R2PerFactor = fvar_mk)
    return(R2_list)
}
```

Function to plot R2: 'plotR2'
```{r, echo=F}
library(scales)
plotR2 <- function(fvar_mk, title="") { 
  
      
      # Sort factors
      fvar_mk_df <- reshape2::melt(fvar_mk, varnames=c("factor","view"))
      fvar_mk_df$factor <- factor(fvar_mk_df$factor)
      hc <- hclust(dist(t(fvar_mk)))
      fvar_mk_df$view <- factor(fvar_mk_df$view, levels = colnames(fvar_mk)[hc$order])
      fvar_mk_df$factor <- factor(fvar_mk_df$factor, levels = rownames(fvar_mk))
      
    if(length(unique(fvar_mk_df$value))>2){
      # Plot 1: grid with the variance explained per factor in each view
      hm <- ggplot(fvar_mk_df, aes(view,factor)) + 
        geom_tile(aes(fill=value), color="black") +
        guides(fill=guide_colorbar("R2")) +
        # scale_fill_gradientn(colors=c("violet","gray97","darkblue"), guide="colorbar", limits=c(-0.05,0.25),
                             # values  = rescale(c(-0.05, 0, 0.25))) +
        scale_fill_gradientn(colors=c("gray97","darkblue"), guide="colorbar", limits=c(0,0.25)) +
        ylab("Latent factor") +
        theme(
          # plot.margin = margin(5,5,5,5),
          plot.title = element_text(size=17, hjust=0.5),
          axis.text.x = element_blank(),
          axis.text.y =  element_blank(),
          axis.title.y = element_text(size=15),
          axis.line = element_blank(),
          axis.ticks =  element_blank(),
          panel.background = element_blank()
          )+ 
        guides(fill=guide_colorbar("R2"))
    } else{
      fvar_mk_df$value <- ifelse(fvar_mk_df$value==1, "yes", "no")
       hm <- ggplot(fvar_mk_df, aes(view,factor)) + 
        geom_tile(aes(fill=as.factor(value)), color="black") +
        guides(fill=guide_colorbar("R2")) +
        scale_fill_manual(values=c("gray97","darkblue")) +
        ylab("Latent factor") +
        theme(
          # plot.margin = margin(5,5,5,5),
          plot.title = element_text(size=17, hjust=0.5),
          axis.text.x = element_blank(),
          axis.text.y =  element_blank(),
          axis.title.y = element_text(size=15),
          axis.line = element_blank(),
          axis.ticks =  element_blank(),
          panel.background = element_blank()
          )+ 
        guides(fill=guide_legend("active"))
    }

      
      hm <- hm + ggtitle(title)  
  
        
    hm
}
```

# Comparison of variance explained plots
```{r}
gg_single <- list()
for(i in seq_along(iCluster.out)){

# get R2 and re-order factors by number of active views:  
r2out_iClust <- calculateVarianceExplained4iClust(iCluster.out[[i]], data = simData, orderFactorsbyR2 = T)
r2out_iClust$R2PerFactor <- r2out_iClust$R2PerFactor[order(colSums(c(101,102,103)*t(r2out_iClust$R2PerFactor>0.05))),]
r2out_MOFA <- calculateVarianceExplained(MOFA.out[[i]], plotit = F)
r2out_MOFA$R2PerFactor <- r2out_MOFA$R2PerFactor[order(colSums(c(101,102,103)*t(r2out_MOFA$R2PerFactor>0.05))),]

#only take top X factors for GFA
r2out_GFA <- calculateVarianceExplained(GFA.out[[i]], plotit = F, factors = 1:names(iCluster.out)[i])
r2out_GFA$R2PerFactor <- r2out_GFA$R2PerFactor[order(colSums(c(101,102,103)*t(r2out_GFA$R2PerFactor>0.05))),]

r2out_GFA_fixed <- calculateVarianceExplained(GFA_fixed.out[[i]], plotit = F, factors = 1:names(iCluster.out)[i])
r2out_GFA_fixed$R2PerFactor <- r2out_GFA_fixed$R2PerFactor[order(colSums(c(101,102,103)*t(r2out_GFA_fixed$R2PerFactor>0.05))),]

r2MOFA <- r2out_MOFA$R2PerFactor
colnames(r2MOFA) <- 0:2
r2iCluster <- r2out_iClust$R2PerFactor
colnames(r2iCluster) <- 0:2
r2GFA <- r2out_GFA$R2PerFactor
colnames(r2GFA) <- 0:2
r2GFA_fixed <- r2out_GFA_fixed$R2PerFactor
colnames(r2GFA_fixed) <- 0:2

# get true pattern
truepattern <- t(1/alpha[[i]])
truepattern <- truepattern[order(colSums(c(101,102,103)*t(truepattern>0.05)), decreasing = F),]
colnames(truepattern) <- 0:2

# make plots
gg_iCluster <- plotR2(pmax(r2iCluster,0), title = "iCluster")
gg_MOFA <- plotR2(pmax(r2MOFA,0), title = "MOFA")
gg_GFA <- plotR2(pmax(r2GFA,0), title = "GFA")
gg_GFA_fixed <- plotR2(pmax(r2GFA_fixed,0), title = "GFA (fixed)")
gg_truth <- plotR2(truepattern, title = "ground truth")

# save
if(i==1) gg_legend <- grid.arrange(ggplot(),get_legend(gg_truth), get_legend(gg_MOFA+ guides(fill=guide_legend(bquote(R^2)))),ggplot(), ncol=1, heights=c(1.5,1,1,1.5))
gg_single[[i]] <-grid.arrange(gg_truth+guides(fill=F),
                              gg_MOFA+guides(fill=F) + ylab(""),
                              gg_iCluster+guides(fill=F) + ylab(""),
                              gg_GFA+guides(fill=F) + ylab(""),
                              gg_GFA_fixed+ ylab("") +guides(fill=F),
                              ncol=5, top =paste("k =", names(iCluster.out)[i]))
}
```

# Arrange plot
```{r R2_comparison, fig.width=12, fig.height=7}
plot_grid(plot_grid(gg_single[[1]], gg_single[[2]], nrow=2, labels = c("a","b")), gg_legend, rel_widths=c(5,0.5), label_size=22)
```

#SessionInfo
```{r}
sessionInfo()
```

