---
title: "CLL Overview Figure"
author: "Britta Velten"
date: "7/4/2017"
output:
  BiocStyle::html_document:
    toc: true
---

# Introduction
This script provides an overview of the factors inferred by MOFA on the CLL data set as shown in Figure 2 of the manuscript.

```{r, message=F, warning=F}
library(MOFAtools)
library(dplyr)
library(reshape2)
library(pheatmap)
library(survival)
library(gridExtra)
library(ggplot2)
library(GGally)
library(magrittr)
library(Hmisc)
library(cowplot)
# library(pace)
```

# Get imported model and related data
"out_import_technical.RData" prepared in CLL_import_technical.Rmd
```{r}
#output path for plots and data
plotdir <- "/Users/bvelten/Documents/MOFA/Analysis_Code/MOFA_CLL/figures"
if(!dir.exists(plotdir)) dir.create(plotdir)
outdir <- "/Users/bvelten/Documents/MOFA/Analysis_Code/MOFA_CLL/out"

# load model selected in import_models.Rmd
load(file.path(outdir,"out_import.RData"))

source("plottingUtils.R")
```

# Variance explained per view and factor
```{r}
pdf(file.path(plotdir,"R2.pdf"), height=8, width=8)
r2out <- calculateVarianceExplained(model, orderFactorsbyR2 = T, showtotalR2 = T)
dev.off()
r2out
```


## 2(b) Flipped R2 plot with spaces
```{r}
p <- plotFlippedR2(r2.out)

pdf(file.path(plotdir, "R2_flipped.pdf"), width=12)
print(p)
dev.off()
```




# Lineplot 2(c): Weights on Mutations
For small views: Direct inspection of weight. Here Mutations
```{r}
# Mutation weights on factor 1
pdf(file.path(plotdir,"MutWeights_1.pdf"), width = 4.4, height=3)
showTopWeightsAndColor(model, "Mutations", "1" , nfeatures = 5, abs=T, Features2color = "IGHV", col2highlight = "red", maxL = 1, scalePerView=T)
dev.off()

# Mutation weights on factor 2
# rename del13q14, this is now taken care of in input, only used for old model
MOFAtools::featureNames(model)$Mutations[MOFAtools::featureNames(model)$Mutations=="del13q14_any"] <- "del13q14"

pdf(file.path(plotdir,"MutWeights_2.pdf"), width = 4.4, height=3)
showTopWeightsAndColor(model, "Mutations", "2" , nfeatures = 5, abs=T, Features2color = "trisomy12",
                       maxL = 1, scalePerView=T)
dev.off()
```

# Scatterplot 2(d) of two most important facotrs - bicolor
```{r}
# collect data for plot
df = data.frame(x = Z[, "1"], y = Z[,"2"], 
                trisomy12 = covariates$trisomy12,
                IGHV = covariates$IGHV)

# nice names for tr 12
df$trisomy12[is.na(df$trisomy12)] <- "missing" 
df$trisomy12[df$trisomy12 == "0"] <- "wt" 
df$trisomy12[df$trisomy12 == "1"] <- "tr12" 
df$trisomy12 <- factor(df$trisomy12, levels=c("wt", "tr12", "missing"))

# nice names for IGHV
df$IGHV[is.na(df$IGHV)] <- "missing" 
df$IGHV[df$IGHV == "0"] <- "U" 
df$IGHV[df$IGHV == "1"] <- "M" 
df$IGHV <- factor(df$IGHV, levels=c("U", "M", "missing"))

# nice names for comi
df$IGHV_tr12 <- paste(df$IGHV, df$trisomy12, sep="-CLL, ")
df %<>% mutate(IGHV_tr12= ifelse(grepl("missing", IGHV_tr12), "missing",IGHV_tr12))

# colors and shapes for plot
Paircolorr <- c(RColorBrewer::brewer.pal(6, "Paired")[c(1:2,5:6)], "grey")
Shapes4plot <- c(17,19,17,19, 3)
names(Shapes4plot) <- names(Paircolorr) <- c("U-CLL, wt", "U-CLL, tr12", "M-CLL, wt", "M-CLL, tr12", "missing")
df$IGHV_tr12 <- factor(df$IGHV_tr12, levels=names(Paircolorr))

# make plot
titlesize <- 15
gg <-  ggplot(df, aes(x=x, y=y, col=IGHV_tr12,shape=IGHV_tr12)) + geom_point(size=2.5)  +
      theme(plot.margin = margin(20, 20, 10, 10), 
            axis.text = element_text(size = rel(1), color = "black"), 
            axis.title = element_text(size = titlesize), 
            axis.title.y = element_text(size = rel(1.1), margin = margin(0, 10, 0, 0)), 
            axis.title.x = element_text(size = rel(1.1), margin = margin(10, 0, 0, 0)), 
            axis.line = element_line(color = "black", size = 0.5), 
            axis.ticks = element_line(color = "black", size = 0.5),
            panel.border = element_blank(), 
            panel.grid.major = element_blank(),
            panel.grid.minor = element_blank(), 
            panel.background = element_blank(),
            legend.key = element_rect(fill = "white"),
            legend.text = element_text(size = titlesize),
            legend.title = element_blank(),
            legend.position = "top") +
  scale_shape_manual(name = "Mutational status", labels= names(Shapes4plot), values=Shapes4plot) +
  scale_color_manual(name = "Mutational status", labels= names(Paircolorr), values=Paircolorr) +
   guides(colour = guide_legend(ncol=3), shape = guide_legend(ncol=3)) +
    xlab("Factor 1") + ylab("Factor 2") 

# save to pdf
pdf(file.path(plotdir,"ScatterPlot_IGHV_tr12_bicolor.pdf"), width = 5.5, height = 6, useDingbats=F)
print(gg)
dev.off()
```

# Enrichment Analysis

## Reactome Gene Set
Get gene set annotations
```{r}
# Get reactome gene sets
reactomceGS<-readRDS("~/Documents/MOFA/CLL_MOFA_data/GO/reactome/human_reactome_binary_matrix.rds")
reactomceGSList <- apply(reactomceGS, 1,  function(c) colnames(reactomceGS)[c==1])

#Get describtion of pathways
ReactomePathwayInfo<-readLines("~/Documents/MOFA/CLL_MOFA_data/GO/reactome/ReactomePathwaysMeta.txt")
ReactomePathwayInfo<-do.call(rbind,strsplit(ReactomePathwayInfo, "\t"))
colnames(ReactomePathwayInfo)<-c("id", "description", "species")
ReactomePathwayInfo<-as.data.frame(ReactomePathwayInfo)
ReactomePathwayInfo<-ReactomePathwayInfo[ReactomePathwayInfo$species=="Homo sapiens",]
rownames(reactomceGS) <- ReactomePathwayInfo$description[match(rownames(reactomceGS), ReactomePathwayInfo$id)]
```

```{r}
# use ensIDs
if(!all(grepl("ENS", featureNames(model)$mRNA))) 
  MOFAtools::featureNames(model)$mRNA <- mRNA$ens_id[match(MOFAtools::featureNames(model)$mRNA, mRNA$symbol)]
```

## Parametric approach
```{r}
gsea.outParam <- FeatureSetEnrichmentAnalysis(model, "mRNA", reactomceGS, statistical.test = "cor.adj.parametric",
                                              alpha = 0.1, min.size = 15)
gsea.outParam_nonadj <- FeatureSetEnrichmentAnalysis(model, "mRNA", reactomceGS, statistical.test = "parametric", alpha = 0.01,
                                                    min.size = 15)

```

## Permutation approach
```{r, fig.width= 12}
if(!file.exists("gsea.out_perm.RData")){
gsea.outPerm <- FeatureSetEnrichmentAnalysis(model, "mRNA", reactomceGS, statistical.test = "permutation", alpha = 0.1,
                                             nperm = 1000, min.size = 15)
save(gsea.outPerm, file="gsea.out_perm_10Aug.RData")
} else load("gsea.out_perm.RData")
```

## GSE results
### Inspection
```{r}
sapply(gsea.outPerm$sigPathways, length)
sapply(gsea.outParam$sigPathways, length)
sapply(gsea.outParam_nonadj$sigPathways, length)

hist(gsea.outParam$pval)
hist(gsea.outPerm$pval)
hist(gsea.outParam_nonadj$pval)
```

We use non-adjusted parametric test in the following:

```{r}
# choose test to use 
gsea.out <- gsea.outParam_nonadj
names(gsea.out$sigPathways) <- colnames(Z)[-1]
```

### Manual categories
```{r}
#broader categories
label_pathway <- function(x){
  ifelse(grepl("[s|S]tress|HSF|[S|s]enescence|[T|t]elomer|Attenuation phase", x), "stress_aging",
         ifelse(grepl("egulat|Polymerase", x) & grepl("RNA", x), "RNA_regulation",
                ifelse(grepl("Immun|TCR|Interleukin|IL",x), "ImmuneResponse", "other")))
}
categories <- sapply(rownames(reactomceGS), label_pathway)

# write categories to .csv files
write.csv(file=file.path(plotdir,"stress_aging.csv"),names(categories[which(categories=="stress_aging")]))
write.csv(file=file.path(plotdir,"ImmuneResponse.csv"),names(categories[which(categories=="ImmuneResponse")]))
write.csv(file=file.path(plotdir,"RNA_regulation.csv"),names(categories[which(categories=="RNA_regulation")]))
write.csv(file=file.path(plotdir,"other.csv"),names(categories[which(categories=="other")]))

# color definition
col4Pathways <- c("other"="gray", "cellular stress/senescence"="cyan", "RNA regulation"="navy", "Immune Response"="forestgreen")
nicelabels <- c(other="other", stress_aging="cellular stress/senescence",ImmuneResponse="Immune Response",RNA_regulation="RNA regulation" )
```

### 2(e) Bar plot
```{r}
pathwaysDF <- melt(gsea.out$sigPathways, value.name="pathway")
colnames(pathwaysDF) <- c("pathway", "factor")
pathwaysDF %<>% mutate(type=label_pathway(pathway))
# summarize per factor
pathwaysSummary <- pathwaysDF %>% dplyr::group_by(factor) %>%
  dplyr::summarise(other = sum(type=="other"), stress_aging=sum(type=="stress_aging"),
                   ImmuneResponse = sum(type == "ImmuneResponse"),
                   RNA_regulation = sum(type=="RNA_regulation"))
# add factors without enrichment
pathwaysSummary <- rbind(pathwaysSummary, 
                         data.frame(factor = colnames(Z)[-1][!colnames(Z)[-1] %in% pathwaysSummary$factor],
                                                     other = 0, stress_aging=0, ImmuneResponse=0, RNA_regulation=0))

pathwaysSummary <-  melt(pathwaysSummary, id.vars = "factor", variable.name = "type", value.name = "count")
pathwaysSummary$factor <- factor(pathwaysSummary$factor, levels = 1:10)
pathwaysSummary %<>% mutate(type=as.character(nicelabels[type]))
pathwaysSummary$type <- factor(pathwaysSummary$type, levels=rev(as.character(nicelabels)))

#plot
pdf(file.path(plotdir,"GeneSetsOverview.pdf"), width = 4, height = 6)
ggplot(pathwaysSummary, aes(x=factor, y=count, fill=type)) +
  geom_bar(stat="identity") +coord_flip() + 
  ylab("Enriched gene sets at FDR 1%") +
  xlab("Factor") + #guides(fill=F) +
  scale_fill_manual(values=col4Pathways) +
  theme(legend.position = "bottom") +
  guides(fill=guide_legend(title="",nrow=4))
dev.off()

pdf(file.path(plotdir,"GeneSetsOverview_rot.pdf"), width = 4, height = 4.8)
ggplot(pathwaysSummary, aes(x=factor, y=count, fill=type)) +
  geom_bar(stat="identity") +
  ylab("Enriched gene sets at FDR 1%") +
  xlab("Factor") + #guides(fill=F) +
  scale_fill_manual(values=col4Pathways) +
  theme(legend.position = "top") +
  guides(fill=guide_legend(title="",nrow=2)) #+scale_y_reverse() +
    #scale_x_discrete(position = "top")
dev.off()
```

### Point jitter plot of p-values
```{r}
pval.adj <- gsea.out$pval.adj
pDF <- melt(pval.adj, value.name = "p.adj", varnames = c("pathway", "factor"))
pDF %<>% filter(paste0(pathway, factor) %in% apply(pathwaysDF[,c("pathway", "factor")],1,paste0, collapse=""))
#pDF %<>% filter(p.adj<0.01)

pDF %<>% mutate(type=label_pathway(pathway))
pDF %<>% mutate(type=as.character(nicelabels[type]))
pDF$type <- factor(pDF$type, levels=rev(as.character(nicelabels)))

pdf(file.path(plotdir,"GeneSetsOverview_points.pdf"), width = 4.5, height = 5)
ggplot(pDF, aes(x=as.factor(factor), y=-log10(p.adj), col=type))+geom_jitter(aes(alpha=type),width = 0.1, height = 0, size=2) +
  coord_flip() +
  scale_color_manual(values=col4Pathways) +
  ylab("-log(p)") + xlab("Factor")+
  theme(legend.position = "bottom") +
  guides(col=guide_legend(title="",nrow=2), alpha=F) +
  scale_alpha_manual(values=c(other=0.4, 0.8,0.8,0.8))
dev.off()
```


### Line plots per factor
```{r}
pvalsadj <- gsea.out$pval.adj
NsigP <- sapply(gsea.out$sigPathways, length)
for(i in factorNames(model)[-1]){
if(NsigP[i]>0){
pdf(file.path(plotdir,paste0("GeneSetsPerFactor_",i,".pdf")), width = 12, height = 5+NsigP[i]/4, useDingbats = T)
print(LinePlot_FeatureSetEnrichmentAnalysis(pvalsadj, i, threshold=0.01, max.pathways=20))
dev.off()
}
}
```

```{r}
sessionInfo()
```

