---
title: "Overview of MOFA factors of the CLL data"
author: "Britta Velten"
date: '`r format(Sys.time(), "%d %B, %Y")`'
output:
  BiocStyle::html_document:
    toc: true
---

# Introduction
This script provides an overview of the factors inferred by MOFA on the CLL data set as shown in Figure 2 of the manuscript.  The analysis is based on the model and other objects defined in import_models.Rmd.

```{r, message=F, warning=F}
library(MOFAtools)
library(dplyr)
library(reshape2)
library(gridExtra)
library(ggplot2)
library(GGally)
library(magrittr)
library(cowplot)
options(stringsAsFactors = FALSE)
```

# Get imported model and related data
"out_import.RData" prepared in import_models.Rmds
```{r}
#output path for plots and data
plotdir <- "figures/MOFAfactors_overview/"
if(!dir.exists(plotdir)) dir.create(plotdir)
knitr::opts_chunk$set(fig.path=plotdir, dev=c("png", "pdf"))
outdir <- "out"

# load model selected in import_models.Rmd
load(file.path(outdir,"out_import.RData"))

source("plotting_utils.R")
```

# Variance explained per view and factor
```{r R2_flipped, fig.width=12}
r2out <- calculateVarianceExplained(model,plotit = FALSE)
r2out

p <- plotFlippedR2(r2.out)

print(p)
```



# Weights on Mutations view for the first two factors
For small views: Direct inspection of weight is a good way to annotate factors, e.g. here for Mutations.
```{r MutWeights, fig.width = 4.4, fig.height=3}
# Mutation weights on factor 1
showTopWeightsAndColor(model, "Mutations", "1" , nfeatures = 5,
                       abs=T, Features2color = "IGHV",
                      maxL = 1, scalePerView=T)


# Mutation weights on factor 2
showTopWeightsAndColor(model, "Mutations", "2" , nfeatures = 5,
                       abs=T, Features2color = "trisomy12",
                       maxL = 1, scalePerView=T)
```

# Scatterplot of two most important facotrs - bicolor
```{r ScatterPlot_IGHV_tr12_bicolor, fig.width = 5.5, fig.height = 6}
# collect data for plot
df = data.frame(x = Z[, "1"], y = Z[,"2"], 
                trisomy12 = covariates$trisomy12,
                IGHV = covariates$IGHV)

# nice names for tr 12
df$trisomy12[is.na(df$trisomy12)] <- "missing" 
df$trisomy12[df$trisomy12 == "0"] <- "wt" 
df$trisomy12[df$trisomy12 == "1"] <- "tr12" 
df$trisomy12 <- factor(df$trisomy12, levels=c("wt", "tr12", "missing"))

# nice names for IGHV
df$IGHV[is.na(df$IGHV)] <- "missing" 
df$IGHV[df$IGHV == "0"] <- "U" 
df$IGHV[df$IGHV == "1"] <- "M" 
df$IGHV <- factor(df$IGHV, levels=c("U", "M", "missing"))

# nice names for combi
df$IGHV_tr12 <- paste(df$IGHV, df$trisomy12, sep="-CLL, ")
df %<>% mutate(IGHV_tr12= ifelse(grepl("missing", IGHV_tr12), "missing",IGHV_tr12))

# colors and shapes for plot
Paircolorr <- c(RColorBrewer::brewer.pal(6, "Paired")[c(1:2,5:6)], "grey")
Shapes4plot <- c(17,19,17,19, 3)
names(Shapes4plot) <- names(Paircolorr) <- c("U-CLL, wt", "U-CLL, tr12", "M-CLL, wt", "M-CLL, tr12", "missing")
df$IGHV_tr12 <- factor(df$IGHV_tr12, levels=names(Paircolorr))

# make plot
titlesize <- 15
gg <-  ggplot(df, aes(x=x, y=y, col=IGHV_tr12,shape=IGHV_tr12)) +
  geom_point(size=2.5)  +
  theme(plot.margin = margin(20, 20, 10, 10), 
        axis.text = element_text(size = rel(1), color = "black"),
        axis.title = element_text(size = titlesize), 
        axis.title.y = element_text(size = rel(1.1),
                                        margin = margin(0, 10, 0, 0)), 
        axis.title.x = element_text(size = rel(1.1),
                                        margin = margin(10, 0, 0, 0)), 
        axis.line = element_line(color = "black", size = 0.5), 
        axis.ticks = element_line(color = "black", size = 0.5),
        panel.border = element_blank(), 
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(), 
        panel.background = element_blank(),
        legend.key = element_rect(fill = "white"),
        legend.text = element_text(size = titlesize),
        legend.title = element_blank(),
        legend.position = "top") +
  scale_shape_manual(name = "Mutational status",
                     labels= names(Shapes4plot),
                     values=Shapes4plot) +
  scale_color_manual(name = "Mutational status",
                     labels= names(Paircolorr),
                     values=Paircolorr) +
  guides(colour = guide_legend(ncol=3),
          shape = guide_legend(ncol=3)) +
  xlab("Factor 1") + ylab("Factor 2") 

print(gg)
```

# Enrichment Analysis
## Reactome Gene Set
Get gene set annotations
```{r}
# Get reactome gene sets
data("reactomeGS", package="MOFAtools")
```

```{r}
# use ensIDs instead of gene symbols
if(!all(grepl("ENS", featureNames(model)$mRNA))) 
  MOFAtools::featureNames(model)$mRNA <- mRNA$ens_id[match(MOFAtools::featureNames(model)$mRNA, mRNA$symbol)]
```

## Calculate Enrichment
```{r}
gsea.out <- FeatureSetEnrichmentAnalysis(model, "mRNA",reactomeGS,
                                         statistical.test = "parametric",
                                         alpha = 0.01, min.size = 15)
names(gsea.out$sigPathways) <- colnames(Z)[-1]
```


## Inspect GSE results
### Define broad categories manually
```{r}
#broader categories
label_pathway <- function(x){
  ifelse(grepl("[s|S]tress|HSF|[S|s]enescence|[T|t]elomer|Attenuation phase", x), "stress_aging",
         ifelse(grepl("egulat|Polymerase", x) & grepl("RNA", x), "RNA_regulation",
                ifelse(grepl("Immun|TCR|Interleukin|IL",x), "ImmuneResponse", "other")))
}
categories <- sapply(rownames(reactomeGS), label_pathway)

# write categories to .csv files
write.csv(file=file.path(outdir,"stress_aging.csv"),
          names(categories[which(categories=="stress_aging")]))
write.csv(file=file.path(outdir,"ImmuneResponse.csv"),
          names(categories[which(categories=="ImmuneResponse")]))
write.csv(file=file.path(outdir,"RNA_regulation.csv"),
          names(categories[which(categories=="RNA_regulation")]))
write.csv(file=file.path(outdir,"other.csv"),
          names(categories[which(categories=="other")]))

# color definition
col4Pathways <- c("other"="gray",
                  "cellular stress/senescence"="cyan",
                  "RNA regulation"="navy",
                  "Immune Response"="forestgreen")

nicelabels <- c(other="other",
                stress_aging="cellular stress/senescence",
                ImmuneResponse="Immune Response",
                RNA_regulation="RNA regulation" )
```

### Bar plot of enriched gene sets on each factor
```{r GeneSetsOverview_rot, fig.width = 4, fig.height = 4.8}
# collect results in dataframe
pathwaysDF <- melt(gsea.out$sigPathways, value.name="pathway")
colnames(pathwaysDF) <- c("pathway", "factor")
pathwaysDF %<>% mutate(type=label_pathway(pathway))

# summarize per factor
pathwaysSummary <- pathwaysDF %>% dplyr::group_by(factor) %>%
  dplyr::summarise(other = sum(type=="other"),
                   stress_aging=sum(type=="stress_aging"),
                   ImmuneResponse = sum(type == "ImmuneResponse"),
                   RNA_regulation = sum(type=="RNA_regulation"))

# add factors without enrichment
df_none <- data.frame(factor = colnames(Z)[-1][!colnames(Z)[-1] %in% pathwaysSummary$factor],
                      other = 0, stress_aging=0,
                      ImmuneResponse=0, RNA_regulation=0)
pathwaysSummary <- rbind(pathwaysSummary, df_none)
pathwaysSummary %<>%  melt(id.vars = "factor", variable.name = "type", value.name = "count")
pathwaysSummary$factor <- factor(pathwaysSummary$factor, levels = 1:10)
pathwaysSummary %<>% mutate(type=as.character(nicelabels[type]))
pathwaysSummary$type <- factor(pathwaysSummary$type, levels=rev(as.character(nicelabels)))

ggplot(pathwaysSummary, aes(x=factor, y=count, fill=type)) +
  geom_bar(stat="identity") +
  ylab("Enriched gene sets at FDR 1%") +
  xlab("Factor") +
  scale_fill_manual(values=col4Pathways) +
  theme(legend.position = "top") +
  guides(fill=guide_legend(title="",nrow=2))
```

# SessionInfo
```{r}
sessionInfo()
```

