---
title: "Association of latent factors to time to next treatment"
author: "Britta Velten"
date: '`r format(Sys.time(), "%d %B, %Y")`'
output:
  BiocStyle::html_document:
    toc: true
---

# Introduction
This script analyses the links of the MOFA factors to clinical outcome as shown in Figure 5 of the manuscript. The analysis is based on the model and other objects defined in import_models.Rmd.


```{r, message=F, warning=F}
library(MOFAtools)
library(dplyr)
library(reshape2)
library(pheatmap)
library(survival)
library(gridExtra)
library(ggplot2)
library(GGally)
library(magrittr)
library(Hmisc)
library(glmnet)
library(cowplot)
```


# I/O options
```{r}
#output path for plots and data
plotdir <- "figures/survival/"
if(!dir.exists(plotdir)) dir.create(plotdir)
knitr::opts_chunk$set(fig.path=plotdir, dev=c("png", "pdf"))
outdir <- "out"
```

# Get fitted model and related data
(Prepared in import_models.Rmd)
```{r}
# load model selected in import_models.Rmd
load(file.path(outdir,"out_import.RData"))

# helper function for some plots
source("plotting_utils.R")
```

# Get survival data from Dietrich et al. 2017
The patmeta and covariates objects are constucted in import_models.Rmd on the basis of the pace package.
```{r}
data(patmeta, package="pace")
survivalData <- as.matrix(patmeta[,c("T5","treatedAfter")])
survivalData<- survivalData[!is.na(survivalData[,1]),]
colnames(survivalData)<-c("time", "status")
  
# subset to common patients
commonPats <- intersect(rownames(Z), rownames(survivalData))
Zcommon <- Z[commonPats,]
survivalData <- survivalData[commonPats,]
covariatesCommon <- covariates[commonPats,]
dataCommmon <- lapply(data, function(dd) dd[,commonPats])

# construct survival object
SurvObject <-Surv(survivalData[,1],survivalData[,2])
```

# Fit univariate Cox models for each factor
All covariates are scaled to ensure comparability of hazard ratio.This, however, looses interpretability of HR for 0-1 groups.
```{r, fig.height=12, fig.width=10}
# fit a cox model for latent factors, scaling predictors
pval_LFs<-sapply(which(apply(Z,2,var, na.rm=T)>0), function(i){
    fit <- coxph(SurvObject  ~  scale(Zcommon[,i]))  
    s <- summary(fit) 
    c(p      = s[["coefficients"]][, "Pr(>|z|)"], 
      coef   = s[["coefficients"]][, "exp(coef)"], 
      lower  = s[["conf.int"]][, "lower .95"], 
      higher = s[["conf.int"]][, "upper .95"])
  })
colnames(pval_LFs) <- colnames(Z)[which(apply(Z,2,var, na.rm=T)>0)]

# collect in data frame
df_survival <- as.data.frame(t(pval_LFs))
df_survival$label <- paste(formatC(df_survival$p, digits = 2), sep="")
df_survival$predictor <- factor(paste("Factor",colnames(pval_LFs)), levels=rev(paste("Factor",colnames(pval_LFs))))
df_survival

# re-orient to HR > 1
df_survival %<>% mutate(positive = coef>1)
df_survival %<>% mutate(y = ifelse(positive,coef, 1/coef))
df_survival %<>% mutate(ymin = ifelse(positive,lower, 1/lower))
df_survival %<>% mutate(ymax = ifelse(positive,higher, 1/higher))
```

## Forest Plot
```{r ForestPlotSurvival_TTT, fig.width = 7, fig.height = 12}
# TTT
ggplot(df_survival, aes(x=predictor, y=y,ymin=ymin,ymax=ymax))+
  geom_pointrange( col='#619CFF')+ coord_flip() +
  scale_x_discrete() + ylab("(Positive) Hazard Ratio")+ 
  scale_y_log10(breaks=c(0.75,1,1.5,2,3), limits=c(min(df_survival$ymin)-0.1,3.1)) +
  geom_hline(aes(yintercept=1), linetype="dotted") + 
  geom_text(aes(label=label, y=2.5),size=5, hjust = "left")+
  theme(text =element_text(size=18),
        axis.ticks.y = element_blank(),
        legend.position="bottom", panel.grid =element_blank(),
        panel.background = element_rect(fill="white"),
        strip.text = element_blank(),
        axis.text.y = element_text(size=16),
        axis.text.x = element_text(size=16),
        plot.title = element_text(size=18)) +
  guides(colour=F) +
  xlab("Factor") + scale_color_discrete(drop=FALSE)
```


# Multivariate models for predicting survival
Harrals C-Index is used as performance measure, 5-fold stratified cross-validation.
Both models using PCs and no penalization as well as using all features with a ridge approach are fitted here.
```{r}
  #stratified CV to include same proportion of uncensored cases
  set.seed(1290)
  uncensored <- SurvObject[,"status"]==1
  cv_ix <- dismo::kfold(1:length(commonPats), k=5, by=uncensored)
  
  # Use same number of principal components as predictors as MOFA factors (without the intercept)
  topPC <- ncol(Z) - 1

  # impute missing data values by mean
  data_imputed <- lapply(dataCommmon, function(view) {
    apply(view,1, function(x) { x[is.na(x)] <- mean(x, na.rm=TRUE); return(x)})
  })
  
  # add concatenated data matrix
  data_imputed$all <- Reduce(cbind, data_imputed)
  
  # construct predictor list fo first 'topPC' PCs each
  FeatureList <- lapply(names(data_imputed), function(singleview){
    dat <- data_imputed[[singleview]]
    pc.out <- prcomp(dat)
    pc.out$x[,1:topPC]
  })
  names(FeatureList) <- paste(names(data_imputed),sep="")
  names(FeatureList)[grep("viab",names(FeatureList))] <- "DrugResponse"
  
  # add MOFA factors
  FeatureList$LF = Zcommon[,-1]

  # List of predictors using all features
  FeatureList_all <- data_imputed
  names(FeatureList_all) <- paste(names(data_imputed),"full",sep="_")

  # fit a cox model and predict on left-out fold
  fit <- lapply(unique(cv_ix), function(i){
    
  #fit coxph for reduced (i.e. MOFA facotrs and PCs)
  c = lapply(FeatureList, function(x) coxph(SurvObject[cv_ix!=i,]  ~ as.matrix(x)[cv_ix!=i,]))
  p = mapply(function(x,y) as.matrix(x[cv_ix==i,]) %*% coef(y), FeatureList, c)
      
  # fit cox model with ridge penalty for all
  c_all = lapply(FeatureList_all, function(x) 
  cv.glmnet(as.matrix(x)[cv_ix!=i,], survivalData[cv_ix!=i,], family="cox", alpha=0))
  p_all = mapply(function(x,y) as.matrix(x[cv_ix==i,]) %*% coef(y, y$lambda.min), FeatureList_all, c_all)
  p_all <- do.call(cbind, p_all)      
  colnames(p_all) <- names(FeatureList_all)
  
  # calculate CI    on left-out fold 
  p_joint <- cbind(p, p_all)
  CI=apply(-p_joint,2, rcorr.cens, SurvObject[cv_ix==i])[1,]
  list(CI=CI,c=c, c_all=c_all)
  })

  # get cross-validated CI
  concordanceCV <- sapply(fit, function(l) l$CI)
  colnames(concordanceCV) <- unique(cv_ix)
  df_hc <- melt(concordanceCV, varnames = c("predictor", "cv_idx"), value.name = "CI")
  df_hc$predictor <- ifelse(df_hc$predictor=="LF", "MOFA factors", as.character(df_hc$predictor))

  # calculate summary statistics across folds
  summaryHc <- aggregate(df_hc$CI,
    by = list(predictors = df_hc$predictor), 
    FUN = function(x) c(mean = mean(x), sd = sd(x),
                        n = length(x)))
  summaryHc <- do.call(data.frame, summaryHc)
  summaryHc$se <- summaryHc$x.sd / sqrt(summaryHc$x.n)

  summaryHc
```

## Define colors for barplot
```{r}
cols4survival <- c("#E69F00","#D55E00","#009E73", "#56B4E9",  "#999999","#0072B2")
names(cols4survival) <- c("Methylation", "Mutations","mRNA",  "Drugs", "all", "MOFA factors")
```

## Main Plot 4(a): PCs as predictors
Plot the resulting prediciton performance when principal componenets are included as predictors in a multivariate Cox model.
```{r HarrelsC_TTT}
summaryHc_main <- filter(summaryHc,
                         predictors %in% c(names(data), "all", "MOFA factors"))
limits <- aes(ymax = summaryHc_main$x.mean + summaryHc_main$se,
              ymin = summaryHc_main$x.mean - summaryHc_main$se)
summaryHc_main$predictors <- factor(summaryHc_main$predictors,
                                    levels =c("Methylation", "Mutations","mRNA",  "Drugs",
                                               "all", "MOFA factors"))

# barplot
  ggplot(summaryHc_main, aes(x=predictors, y=x.mean, fill=predictors, group=predictors))+
    geom_bar(stat = "identity") + 
   coord_cartesian(ylim = c(0.5,0.85)) +
  scale_fill_manual(values=cols4survival, labels=names(cols4survival)) + 
  geom_errorbar(limits, position = "dodge", width = 0.25)+
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size=18),
        axis.text.y = element_text(size=18),
         axis.title.y = element_text(size=18),
        strip.background = element_rect(fill="white"),
        strip.text = element_blank(),
        legend.position= "none",
        plot.title = element_text(size=20)) +
  ylab("Harrell's C-index") + xlab("")
```

## Supplement S14: All features as predictors
Plot the resulting prediciton performance when all features are included as predictors in a multivariate Cox model with ridge penalty.
```{r HarrelsC_TTT_supp}
summaryHc_supp <- filter(summaryHc,
                         predictors %in% c(paste(c(names(data), "all"), "full", sep="_"), "MOFA factors"))#, "factors 1,7,8"))
limits <- aes(ymax = summaryHc_supp$x.mean + summaryHc_supp$se,
              ymin = summaryHc_supp$x.mean - summaryHc_supp$se)
summaryHc_supp$predictors <- factor(sapply(summaryHc_supp$predictors, function(nm) sub("_full","",nm)),
                                    levels =c("Methylation", "Mutations","mRNA",  "Drugs",
                                               "all", "MOFA factors"))

# barplot
  ggplot(summaryHc_supp, aes(x=predictors, y=x.mean, fill=predictors, group=predictors))+
    geom_bar(stat = "identity") + 
   coord_cartesian(ylim = c(0.5,0.85)) +
  scale_fill_manual(values=cols4survival, labels=names(cols4survival)) + 
  geom_errorbar(limits, position = "dodge", width = 0.25)+
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        strip.background = element_rect(fill="white"),
        strip.text = element_blank(),
        legend.position= "none",
        plot.title = element_text(size=20)) +
  ylab("Harrell's C-index") + xlab("")
```


# Kaplan-Meier plots
To visualize the association of the factor to clinical outcome. Samples are split into two groups based on the continious factors and a Kaplan Meier plot is made for those groups. Note that this is mainly for visualisation purposes, usually the continious factor should be considered as above in the Cox models.
```{r KaplanMeier, fig.width=10.5, fig.height=3.5}
  # calculate KM for all LFs
  glist <- lapply(as.character(1:(ncol(Z)-1)), function(lfno){
  
  #determine optimal cut-point and classify samples accordingly
  df <- data.frame(time=survivalData[,1], event = survivalData[,2], Zcommon)
  cut <- survminer::surv_cutpoint(df, variables=paste("X", lfno, sep=""))
  df$FactorCluster <- Zcommon[,lfno] > cut$cutpoint$cutpoint
  
  # due to rotational invariance of factors just use same colors for upper and lower group in the KM plot 
  # (irrescpecitve of which end of the factor they belong to)
  if(lfno==7) df$FactorCluster <- Zcommon[,lfno] < cut$cutpoint$cutpoint

  # fit survival model for the given factor
  fit <- survfit(Surv(time, event) ~ FactorCluster, df)
  ggLF <- survminer::ggsurvplot(fit, data =df,
                        conf.int = TRUE,
                        pval = TRUE,
                        fun = function(y) y*100,
                        legend = "none",
                        legend.labs = c(paste("low LF", lfno), paste("high LF", lfno)),
                        xlab = "Time to treatment",
                        ylab=ifelse(lfno==1, "Survival probability (%)", ""),
                        title= paste("Factor", lfno)
                        )
  ggLF$plot
 })
 
 # make joint plot of significant factors:
    grid.arrange(glist[[1]], glist[[7]], glist[[8]], ncol=3) 
```

# Connections to clinical covariates

## Load clinical covariates
```{r}
covariates_mofa <- covariates

# Get patient IDs transaltion form P to H numbers
load("~/Documents/cll/var/encPatientID_160218.RData")
rownames(encPatientID) <- encPatientID$PatientID

# doubling times (only available for 90 patients)
load("~/Documents/CLL/OtherData/DoublingTime_070816.RData")
LDT <- as.data.frame(LDT)
rownames(LDT) <- LDT$patID
rownames(LDT) <- encPatientID[rownames(LDT),]$PatientID2
doublingtimes <- LDT[rownames(Z),]$doubling.time

# Add Mutations and LDT to covariates
covariates <- cbind(covariates_mofa[rownames(Z),],
                    t(model@TrainData$Mutations)[rownames(Z),!colnames(t(model@TrainData$Mutations))%in% colnames(covariates_mofa)],
                    doubling.times =doublingtimes)

# Clinical markers to compare
biomarkers <- c("IGHV", "TP53", "age", "sex",
                "MethylationCluster",
                 "pretreated",
                 "doubling times")

# numeric code for IGHV and MC
covariates[,"IGHV"] <- ifelse(covariates[,"IGHV"]=="U",0,1)
covariates[,"MethylationCluster"] <- ifelse(covariates[,"MethylationCluster"]=="LP",0,
                                            ifelse(covariates[,"MethylationCluster"]=="IP",0.5,1))

#nice names
colnames(covariates) <- sapply(colnames(covariates), function(s) ifelse(s=="IC50beforeTreatment","pretreated",s) )
colnames(covariates) <- sapply(colnames(covariates), function(s) ifelse(s=="doubling.times","doubling times",s) )

# get biomarkers
all(biomarkers %in% colnames(covariates))
covariates <- covariates[,biomarkers]
```

## Marginal Cox models
All covariates are scaled to ensure comparability of hazard ratio.This, however, looses interpretability of HR for 0-1 groups.
```{r}
  # subset to patients having the clinical markers
  commonPats <- intersect(commonPats, rownames(covariates)[!apply(covariates,1, function(pat) any(is.na(pat)))])
  print(length(commonPats))
  print(paste("uncensored:",sum(survivalData[,2])))
  Zcommon <- Z[commonPats,]
  covariatesCommon <- covariates[commonPats,]
  survivalData <- survivalData[commonPats,]
    
  # construct survival object
  SurvObject <-Surv(survivalData[,1],survivalData[,2])
  
  # fit a cox model for latent factors, scaling predictors
  pval_LFs <- sapply(which(apply(Z,2,var, na.rm=T)>0), function(i){
    fit <- coxph(SurvObject  ~  scale(Zcommon[,i]))  
    s <- summary(fit) 
    c(p      = s[["coefficients"]][, "Pr(>|z|)"], 
      coef   = s[["coefficients"]][, "exp(coef)"], 
      lower  = s[["conf.int"]][, "lower .95"], 
      higher = s[["conf.int"]][, "upper .95"])
  })
  colnames(pval_LFs) <- paste("Factor",colnames(Z)[which(apply(Z,2,var, na.rm=T)>0)])

  # fit a cox model for other relevant biomarkers, scaling all predictor
  pvals_Cov<-sapply(biomarkers, function(cnm){
    fit <- coxph(SurvObject  ~  scale(covariatesCommon[,cnm])) 
    s <- summary(fit) 
    c(p      = s[["coefficients"]][, "Pr(>|z|)"], 
      coef   = s[["coefficients"]][, "exp(coef)"], 
      lower  = s[["conf.int"]][, "lower .95"], 
      higher = s[["conf.int"]][, "upper .95"])
  })

  # collect in data frame
  pvalsSurv<-cbind(pval_LFs,pvals_Cov)
  df_surv <- as.data.frame(t(pvalsSurv))
  df_surv$predictor <- rownames(df_surv)
  types <- list(latentFactors = colnames(pval_LFs), 
                mutations = biomarkers[biomarkers%in% rownames(model@TrainData$Mutations)], 
                demographics  = biomarkers[grepl("sex|age", biomarkers)],
                clinical = biomarkers[grepl("doubling|pretreated", biomarkers)],
                other = biomarkers[grepl("MethylationCluster", biomarkers)])

  df_surv$predictorType <- with(df_surv, ifelse(predictor %in% types$latentFactors, "MOFA factors",
                              ifelse(predictor %in% types$mutations, "genetic",
                                     ifelse(predictor %in% types$clinical, "clinical",
                                            ifelse(predictor %in% types$demographics, "demographic", "other")))))
  df_surv$predictor <- factor(df_surv$predictor, levels = rev(Reduce(c, types)))
  df_surv$survType <- "Time to next treatment"
  df_surv$label <- paste(formatC(df_surv$p, digits = 2), sep="")
  df_surv
```

Make Forest plot
```{r forestplot_clinical}
col4Types <- c(cols4survival["MOFA factors"], cols4survival["Mutations"],
              "black", "darkgreen", cols4survival["Methylation"])
names(col4Types) <- c('MOFA factors', "genetic", "demographic", "clinical", "other")
gg_forest <- ggplot(filter(df_surv, !predictor %in% paste("Factor",1:10)| p < 0.01), aes(x=predictor,y=coef,ymin=lower,ymax=higher, col=predictorType))+
  geom_pointrange()+ coord_flip() +
  scale_x_discrete() + ylab("Hazard Ratio")+
  scale_y_log10(breaks=c(0.1,0.3,1,3), limits=c(min(df_surv$lower),5)) +
  geom_hline(aes(yintercept=1), linetype="dotted") + facet_wrap(~survType) +
  geom_text(aes(label=label, y=4),size=4.5, vjust = 0)+
  theme(text =element_text(size=16),
        axis.ticks.y = element_blank(),
        legend.position="bottom", panel.grid =element_blank(),
        panel.background = element_rect(fill="white"),
        strip.text = element_blank(),
        axis.text.y = element_text(size=16),
        plot.title = element_text(size=18)) +
  guides(colour=guide_legend("predictor type", ncol=2)) +
  xlab("") + scale_color_manual(values=col4Types)
gg_forest
```


## C-Index
Harrals C-Index is used as performance measure, 5-fold stratified cross-validation.
```{r}
  #stratified CV to include same proportion of uncensored cases
  set.seed(1290)
  uncensored <- SurvObject[,"status"]==1
  cv_ix <- dismo::kfold(1:length(commonPats), k=5, by=uncensored)
  
  # add covariates from forest plot
  FeatureList <- list()
  FeatureList$factors = Zcommon[,c("1","7","8")]
  FeatureList$clinical <- covariatesCommon[colnames(covariatesCommon) %in% types$clinical]
  FeatureList$demographic <- covariatesCommon[colnames(covariatesCommon) %in% types$demographics]
  FeatureList$other <- covariatesCommon[colnames(covariatesCommon) %in% types$other]
  FeatureList$genetic <- covariatesCommon[colnames(covariatesCommon) %in% types$mutations]

  # fit a cox model and predict on left-out fold
  fit <- lapply(unique(cv_ix), function(i){
      c = lapply(FeatureList, function(x) coxph(SurvObject[cv_ix!=i,]  ~ as.matrix(x)[cv_ix!=i,]))
      p = mapply(function(x,y) as.matrix(x[cv_ix==i,]) %*% coef(y), FeatureList, c)
      CI=apply(-p,2, rcorr.cens, SurvObject[cv_ix==i])[1,]
      list(CI=CI,c=c)
  })

    # get cross-validated CI
    concordanceCV <- sapply(fit, function(l) l$CI)
    colnames(concordanceCV) <- unique(cv_ix)
    df_hc <- melt(concordanceCV, varnames = c("predictor", "cv_idx"), value.name = "CI")
    df_hc$survType <-"Time to next treatment"
    df_hc$predictor <- ifelse(df_hc$predictor=="factors", "MOFA factors", as.character(df_hc$predictor))
    df_hc


# calculate summary statistics across folds
summaryHc <- aggregate(df_hc$CI,
    by = list(predictors = df_hc$predictor, survType=df_hc$survType), 
    FUN = function(x) c(mean = mean(x), sd = sd(x),
                        n = length(x)))
summaryHc <- do.call(data.frame, summaryHc)
summaryHc$se <- summaryHc$x.sd / sqrt(summaryHc$x.n)

summaryHc
```

Make barplot
```{r barplot_clinical}
limits <- aes(ymax = summaryHc$x.mean + summaryHc$se,
              ymin = summaryHc$x.mean - summaryHc$se)

# barplot
  gg_bar <- ggplot(summaryHc, aes(x=predictors, y=x.mean, fill=predictors, group=predictors))+
    geom_bar(stat = "identity") + 
   coord_cartesian(ylim = c(0.5,0.85)) +
  # ggtitle(paste("Prediction of time to treatment"))+
  scale_fill_manual(values=col4Types) + 
  geom_errorbar(limits, position = "dodge", width = 0.25)+
  theme(axis.text.x = element_text(angle = 90, vjust=0.5,hjust = 1, size=16, lineheight=-2),
        axis.text.y = element_text(size=16),
         axis.title.y = element_text(size=16),
        strip.background = element_rect(fill="white"),
        strip.text = element_blank(),
        legend.position= "none",
        plot.title = element_text(size=18)) +
  ylab("Harrell's C-index") + xlab("")
gg_bar
```

## Arrange plots
```{r TTT_clinical, fig.width=10, fig.height=7}
plot_grid(gg_forest, gg_bar, labels = c("a","b"), rel_widths=c(1.5,1), label_size = 25)
```



# SessionInfo
```{r}
sessionInfo()
```

