---
title: "Association of latent factors to time to next treatment"
author: "Britta Velten"
date: "04/07/2017"
output:
  BiocStyle::html_document:
    toc: true
---

# Introduction
This script analyses the links of the MOFA factors to clinical outcome as shown in Figure 5 of the manuscript. The analysis is based on the model and other objects defined in import_models.Rmd.


```{r, message=F, warning=F}
library(MOFAtools)
library(dplyr)
library(reshape2)
library(pheatmap)
library(survival)
library(gridExtra)
library(ggplot2)
library(GGally)
library(magrittr)
library(Hmisc)
library(glmnet)
library(cowplot)
```

# Get imported model and related data
Prepared in CLL_import_technical.Rmd
```{r}
#output path for plots and data
plotdir <- "figures/survival"
if(!dir.exists(plotdir)) dir.create(plotdir)
outdir <- "out"

# load model selected in import_models.Rmd
load(file.path(outdir,"out_import.RData"))
```

# Get survival data from Dietrich et al
The patmeta and covariates objects are constucted in import_models.Rmd on the basis of the pace package.
```{r}
data(patmeta, package="pace")
survivalData <- as.matrix(patmeta[,c("T5","treatedAfter")])
survivalData<- survivalData[!is.na(survivalData[,1]),]
colnames(survivalData)<-c("time", "status")
  
# subset to common patients
commonPats <- intersect(rownames(Z), rownames(survivalData))
Zcommon <- Z[commonPats,]
survivalData <- survivalData[commonPats,]
covariatesCommon <- covariates[commonPats,]
dataCommmon <- lapply(data, function(dd) dd[,commonPats])

# construct survival object
SurvObject <-Surv(survivalData[,1],survivalData[,2])
```

# Fit univariate Cox models for each factor
All covariates are scaled to ensure comparability of hazard ratio.This, however, looses interpretability of HR for 0-1 groups.
```{r, fig.height=12, fig.width=10}
# fit a cox model for latent factors, scaling predictors
pval_LFs<-sapply(which(apply(Z,2,var, na.rm=T)>0), function(i){
    fit <- coxph(SurvObject  ~  scale(Zcommon[,i]))  
    s <- summary(fit) 
    c(p      = s[["coefficients"]][, "Pr(>|z|)"], 
      coef   = s[["coefficients"]][, "exp(coef)"], 
      lower  = s[["conf.int"]][, "lower .95"], 
      higher = s[["conf.int"]][, "upper .95"])
  })
colnames(pval_LFs) <- colnames(Z)[which(apply(Z,2,var, na.rm=T)>0)]

# collect in data frame
df_survival <- as.data.frame(t(pval_LFs))
df_survival$label <- paste(formatC(df_survival$p, digits = 2), sep="")
df_survival$predictor <- factor(paste("Factor",colnames(pval_LFs)), levels=rev(paste("Factor",colnames(pval_LFs))))
df_survival

# re-orient to HR > 1
df_survival %<>% mutate(positive = coef>1)
df_survival %<>% mutate(y = ifelse(positive,coef, 1/coef))
df_survival %<>% mutate(ymin = ifelse(positive,lower, 1/lower))
df_survival %<>% mutate(ymax = ifelse(positive,higher, 1/higher))
```

## Forest Plot
```{r, fig.width = 7, fig.height = 12}
# pdf(file.path(plotdir, "ForestPlotSurvival_TTT.pdf"), width = 7, height = 12)
# TTT
ggplot(df_survival, aes(x=predictor, y=y,ymin=ymin,ymax=ymax))+
  geom_pointrange( col='#619CFF')+ coord_flip() +
  scale_x_discrete() + ylab("(Positive) Hazard Ratio")+ 
  scale_y_log10(breaks=c(0.75,1,1.5,2,3), limits=c(min(df_survival$ymin)-0.1,3.1)) +
  geom_hline(aes(yintercept=1), linetype="dotted") + 
  geom_text(aes(label=label, y=2.5),size=5, hjust = "left")+
  theme(text =element_text(size=18),
        axis.ticks.y = element_blank(),
        legend.position="bottom", panel.grid =element_blank(),
        panel.background = element_rect(fill="white"),
        strip.text = element_blank(),
        axis.text.y = element_text(size=16),
        axis.text.x = element_text(size=16),
        plot.title = element_text(size=18)) +
  guides(colour=F) +
  xlab("Factor") + scale_color_discrete(drop=FALSE)
# dev.off()
```


# Multivariate models for predicting survival
Harrals C-Index is used as performance measure, 5-fold stratified cross-validation.
Both models using PCs and no penalization as well as using all features with a ridge approach are fitted here.
```{r}
  #stratified CV to include same proportion of uncensored cases
  set.seed(1290)
  uncensored <- SurvObject[,"status"]==1
  cv_ix <- dismo::kfold(1:length(commonPats), k=5, by=uncensored)
  
  # Use same number of principal components as predictors as MOFA factors (without the intercept)
  topPC <- ncol(Z) - 1

  # impute missing data values by mean
  data_imputed <- lapply(dataCommmon, function(view) {
    apply(view,1, function(x) { x[is.na(x)] <- mean(x, na.rm=TRUE); return(x)})
  })
  
  # add concatenated data matrix
  data_imputed$all <- Reduce(cbind, data_imputed)
  
  # construct predictor list fo first 'topPC' PCs each
  FeatureList <- lapply(names(data_imputed), function(singleview){
    dat <- data_imputed[[singleview]]
    pc.out <- prcomp(dat)
    pc.out$x[,1:topPC]
  })
  names(FeatureList) <- paste(names(data_imputed),sep="")
  names(FeatureList)[grep("viab",names(FeatureList))] <- "DrugResponse"
  
  # add MOFA factors
  FeatureList$LF = Zcommon[,-1]

  # List of predictors using all features
  FeatureList_all <- data_imputed
  names(FeatureList_all) <- paste(names(data_imputed),"full",sep="_")

  # fit a cox model and predict on left-out fold
  fit <- lapply(unique(cv_ix), function(i){
    
  #fit coxph for reduced (i.e. MOFA facotrs and PCs)
  c = lapply(FeatureList, function(x) coxph(SurvObject[cv_ix!=i,]  ~ as.matrix(x)[cv_ix!=i,]))
  p = mapply(function(x,y) as.matrix(x[cv_ix==i,]) %*% coef(y), FeatureList, c)
      
  # fit cox model with ridge penalty for all
  c_all = lapply(FeatureList_all, function(x) 
  cv.glmnet(as.matrix(x)[cv_ix!=i,], survivalData[cv_ix!=i,], family="cox", alpha=0))
  p_all = mapply(function(x,y) as.matrix(x[cv_ix==i,]) %*% coef(y, y$lambda.min), FeatureList_all, c_all)
  p_all <- do.call(cbind, p_all)      
  colnames(p_all) <- names(FeatureList_all)
  
  # calculate CI    on left-out fold 
  p_joint <- cbind(p, p_all)
  CI=apply(-p_joint,2, rcorr.cens, SurvObject[cv_ix==i])[1,]
  list(CI=CI,c=c, c_all=c_all)
  })

  # get cross-validated CI
  concordanceCV <- sapply(fit, function(l) l$CI)
  colnames(concordanceCV) <- unique(cv_ix)
  df_hc <- melt(concordanceCV, varnames = c("predictor", "cv_idx"), value.name = "CI")
  df_hc$predictor <- ifelse(df_hc$predictor=="LF", "MOFA factors", as.character(df_hc$predictor))

  # calculate summary statistics across folds
  summaryHc <- aggregate(df_hc$CI,
    by = list(predictors = df_hc$predictor), 
    FUN = function(x) c(mean = mean(x), sd = sd(x),
                        n = length(x)))
  summaryHc <- do.call(data.frame, summaryHc)
  summaryHc$se <- summaryHc$x.sd / sqrt(summaryHc$x.n)

  summaryHc
```

## Define colors for barplot
```{r}
cols4survival <- c("#E69F00","#D55E00","#009E73", "#56B4E9",  "#999999","#0072B2")
names(cols4survival) <- c("Methylation", "Mutations","mRNA",  "Drugs", "all", "MOFA factors")
```

## Main Plot 4(a): PCs as predictors
Plot the resulting prediciton performance when principal componenets are included as predictors in a multivariate Cox model.
```{r}
summaryHc_main <- filter(summaryHc,
                         predictors %in% c(names(data), "all", "MOFA factors"))
limits <- aes(ymax = summaryHc_main$x.mean + summaryHc_main$se,
              ymin = summaryHc_main$x.mean - summaryHc_main$se)
summaryHc_main$predictors <- factor(summaryHc_main$predictors,
                                    levels =c("Methylation", "Mutations","mRNA",  "Drugs",
                                               "all", "MOFA factors"))

# barplot
# pdf(file.path(plotdir, "HarrelsC_TTT.pdf"))
  ggplot(summaryHc_main, aes(x=predictors, y=x.mean, fill=predictors, group=predictors))+
    geom_bar(stat = "identity") + 
   coord_cartesian(ylim = c(0.5,0.85)) +
  scale_fill_manual(values=cols4survival, labels=names(cols4survival)) + 
  geom_errorbar(limits, position = "dodge", width = 0.25)+
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size=18),
        axis.text.y = element_text(size=18),
         axis.title.y = element_text(size=18),
        strip.background = element_rect(fill="white"),
        strip.text = element_blank(),
        legend.position= "none",
        plot.title = element_text(size=20)) +
  ylab("Harrell's C-index") + xlab("")
# dev.off()
```

## Supplement S14: All features as predictors
Plot the resulting prediciton performance when all features are included as predictors in a multivariate Cox model with ridge penalty.
```{r}
summaryHc_supp <- filter(summaryHc,
                         predictors %in% c(paste(c(names(data), "all"), "full", sep="_"), "MOFA factors"))#, "factors 1,7,8"))
limits <- aes(ymax = summaryHc_supp$x.mean + summaryHc_supp$se,
              ymin = summaryHc_supp$x.mean - summaryHc_supp$se)
summaryHc_supp$predictors <- factor(sapply(summaryHc_supp$predictors, function(nm) sub("_full","",nm)),
                                    levels =c("Methylation", "Mutations","mRNA",  "Drugs",
                                               "all", "MOFA factors"))

# barplot
# pdf(file.path(plotdir, "HarrelsC_TTT_supp.pdf"))
  ggplot(summaryHc_supp, aes(x=predictors, y=x.mean, fill=predictors, group=predictors))+
    geom_bar(stat = "identity") + 
   coord_cartesian(ylim = c(0.5,0.85)) +
  scale_fill_manual(values=cols4survival, labels=names(cols4survival)) + 
  geom_errorbar(limits, position = "dodge", width = 0.25)+
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        strip.background = element_rect(fill="white"),
        strip.text = element_blank(),
        legend.position= "none",
        plot.title = element_text(size=20)) +
  ylab("Harrell's C-index") + xlab("")
# dev.off()
```


# Kaplan-Meier plots
To visualize the association of the factor to clinical outcome. Samples are split into two groups based on the continious factors and a Kaplan Meier plot is made for those groups. Note that this is mainly for visualisation purposes, usually the continious factor should be considered as above in the Cox models.
```{r, fig.width=10.5, fig.height=3.5}
  # calculate KM for all LFs
  glist <- lapply(as.character(1:(ncol(Z)-1)), function(lfno){
  
  #determine optimal cut-point and classify samples accordingly
  df <- data.frame(time=survivalData[,1], event = survivalData[,2], Zcommon)
  cut <- survminer::surv_cutpoint(df, variables=paste("X", lfno, sep=""))
  df$FactorCluster <- Zcommon[,lfno] > cut$cutpoint$cutpoint
  
  # due to rotational invariance of factors just use same colors for upper and lower group in the KM plot 
  # (irrescpecitve of which end of the factor they belong to)
  if(lfno==7) df$FactorCluster <- Zcommon[,lfno] < cut$cutpoint$cutpoint

  # fit survival model for the given factor
  fit <- survfit(Surv(time, event) ~ FactorCluster, df)
  ggLF <- survminer::ggsurvplot(fit, data =df,
                        conf.int = TRUE,
                        pval = TRUE,
                        fun = function(y) y*100,
                        legend = "none",
                        legend.labs = c(paste("low LF", lfno), paste("high LF", lfno)),
                        xlab = "Time to treatment",
                        ylab=ifelse(lfno==1, "Survival probability (%)", ""),
                        title= paste("Factor", lfno)
                        )
  ggLF$plot
 })
 
 # make joint plot of significant factors:
  # pdf(file.path(plotdir, "KaplanMeier.pdf"), width=10.5, height=3.5)
    grid.arrange(glist[[1]], glist[[7]], glist[[8]], ncol=3) 
  # dev.off()
```


# SessionInfo
```{r}
sessionInfo()
```

