---
title: "Explore factors 8 and 7"
author: "Britta Velten"
date: "11/8/2017"
output:
  BiocStyle::html_document:
    toc: true
---

```{r, message=F, warning=F}
setwd("/Users/bvelten/Documents/MOFA/CLL_MOFA_data/Analysis/code4manuscript/")
devtools::load_all("/Users/bvelten/Documents/MOFA/MOFApackage/scGFA/MOFAtools/")
# library(MOFAtools)
library(dplyr)
library(reshape2)
library(pheatmap)
library(survival)
library(gridExtra)
library(ggplot2)
library(GGally)
library(magrittr)
library(Hmisc)
library(cowplot)
# library(pace)
```

```{r}
knitr::opts_chunk$set(fig.height = 10)
```

# Get imported model and related data
Prepared in CLL_import_technical.Rmd
```{r}
load("../code4manuscript/out_import_technical_flipped.RData")
model@ModelOpts$learnIntercept <- TRUE
#output path for plots
plotdir <- "~/Documents/MOFA/CLL_MOFA_data/Analysis/exploratory/figures"
if(!dir.exists(plotdir)) dir.create(plotdir)
```

# Which factor is active where?
```{r}
r2 <- calculateVarianceExplained(model)
pdf(file.path(plotdir,"Factor78_R2.pdf"))
r2df <- melt(r2$R2PerFactor[c(7,8),], varnames = c("Factor", "Omic"))
r2df$Factor <- as.factor(r2df$Factor )
gg<- ggplot(r2df, aes(x=Omic, y=value, group=Factor, fill=Factor)) +geom_bar(stat="identity", position = "dodge") +ylab("R2")+ scale_fill_manual(values=c("7"="darkgreen", "8"="navy")) +xlab("")
print(gg)
dev.off()
```

# Association to pretreatment
```{r}
pdf(file.path(plotdir,"Scatterplot_pretreatment.pdf"))
FactorsScatterPlot(model, c("7","8"), color_by = covariates$IC50beforeTreatment)
dev.off()
```

```{r}
pdf(file.path(plotdir,"Factor78_pretreatment.pdf"))
df <- data.frame(pretreated=covariates$IC50beforeTreatment, Factor7= getFactors(model, factors="7"),Factor7= getFactors(model, factors="8"))

df <- filter(df, !is.na(pretreated))
colnames(df) <- c("pretreated", "Factor 7", "Factor 8")
df <- melt(df, variable.name = "factor")

gg <- ggplot(df, aes(x=pretreated, y=value, fill= pretreated)) + geom_violin() +facet_wrap(~factor) + scale_fill_manual(values=c("FALSE"="cornflowerblue", "TRUE"="orange4")) +geom_jitter(alpha=0.5, width=0.05) + ylab("Factor value") 
print(gg)

t.test(value~pretreated, filter(df, factor=="Factor 7") )
t.test(value~pretreated, filter(df, factor=="Factor 8") )
dev.off()
```


# Acitvity in Drug view
```{r}
pdf(file.path(plotdir,"drugs_7.pdf"))
plotTopWeights(model, "Drugs", "7", nfeatures = 15)
dev.off()

pdf(file.path(plotdir,"drugs_8.pdf"))
plotTopWeights(model, "Drugs", "8", nfeatures = 15)
dev.off()
```
Both factors have high eights on rotenone.

# Activity in Mutation View
```{r}
pdf(file.path(plotdir,"mut_7.pdf"))
plotTopWeights(model, "Mutations", "7", nfeatures = 5)
dev.off()
```

# Activity in mRNA View
```{r}
pdf(file.path(plotdir,"mRNA_7.pdf"))
plotTopWeights(model, "mRNA", "7", nfeatures = 10)
dev.off()

pdf(file.path(plotdir,"mRNA_8.pdf"))
plotTopWeights(model, "mRNA", "8", nfeatures = 10)
dev.off()
```

## Reactome Gene Set
Get gene set annotations
```{r}
# Get reactome gene sets
reactomceGS<-readRDS("~/Documents/MOFA/CLL_MOFA_data/GO/reactome/human_reactome_binary_matrix.rds")
reactomceGSList <- apply(reactomceGS, 1,  function(c) colnames(reactomceGS)[c==1])

#Get describtion of pathways
ReactomePathwayInfo<-readLines("~/Documents/MOFA/CLL_MOFA_data/GO/reactome/ReactomePathwaysMeta.txt")
ReactomePathwayInfo<-do.call(rbind,strsplit(ReactomePathwayInfo, "\t"))
colnames(ReactomePathwayInfo)<-c("id", "description", "species")
ReactomePathwayInfo<-as.data.frame(ReactomePathwayInfo)
ReactomePathwayInfo<-ReactomePathwayInfo[ReactomePathwayInfo$species=="Homo sapiens",]
rownames(reactomceGS) <- ReactomePathwayInfo$description[match(rownames(reactomceGS), ReactomePathwayInfo$id)]
```

```{r}
pdf("mRNA_loadings_7.pdf")
# use ensIDs
if(!all(grepl("ENS", featureNames(model)$mRNA))) 
  MOFAtools::featureNames(model)$mRNA <- mRNA$ens_id[match(MOFAtools::featureNames(model)$mRNA, mRNA$symbol)]

gsea.outParam_nonadj <- FeatureSetEnrichmentAnalysis(model, "mRNA", reactomceGS, statistical.test = "parametric", alpha = 0.2, min.size = 15, factors = c("7","8"))
Heatmap_FeatureSetEnrichmentAnalysis(gsea.outParam_nonadj, threshold=0.2)

pdf(file.path(plotdir,"mRNAenr_8.pdf"))
LinePlot_FeatureSetEnrichmentAnalysis(gsea.outParam_nonadj, factor = "8", threshold=0.2)
dev.off()
```


```{r}
knitr::opts_chunk$set(eval=FALSE)
```

But no strong association
```{r}
scatterPlot(model, c("8", "7"),
            color_by = as.factor(covariates$del17p13))
df <- as.data.frame(cbind(Z=Z, covariates, rotenone = colMeans(model@TrainData$Drugs[grepl("rotenone", rownames(model@TrainData$Drugs)),])))
df <- filter(df, !is.na(TP53) & !is.na(IGHV) & !is.na(IC50beforeTreatment))
gg1 <- ggplot(df, aes(y=Z.7, x=IC50beforeTreatment, group=IC50beforeTreatment)) +geom_boxplot() +facet_wrap(paste("TP53", TP53) ~ paste("IGHV", IGHV)) +geom_point()
gg2 <- ggplot(df, aes(y=rotenone, x=IC50beforeTreatment, group=IC50beforeTreatment)) +geom_boxplot() +facet_wrap(paste("TP53", TP53) ~ paste("IGHV", IGHV)) +geom_point()
gg3 <- ggplot(df, aes(y=Z.8, x=IC50beforeTreatment, group=IC50beforeTreatment)) +geom_boxplot() +facet_wrap(paste("TP53", TP53) ~ paste("IGHV", IGHV))
grid.arrange(gg1,gg2, ncol=2)

scatterPlot(model, c("8", "7"),
            color_by = as.factor(covariates$IC50beforeTreatment))
```

#Test for association to covariates
All facotrs significantly associated to survival are also associated to IC50BeforeTreatment!
```{r}
# calculate significance in a linear model and correct for multiple testing per latent factor
pvals_Covariates_LF<-  sapply(colnames(Z)[-1], function(lfidx){
    pvalPerLF<-sapply(colnames(covariates), function(mutName){
      s <- summary(lm(Z[,lfidx] ~ covariates[,mutName]))
      s
      f<-s$fstatistic
      p <- pf(f[1],f[2],f[3],lower.tail=F)
      as.numeric(p)
  })
  })
pvalsadj_Covariates_LF <- matrix(p.adjust(pvals_Covariates_LF, method="BH"), nrow= nrow(pvals_Covariates_LF))
colnames(pvalsadj_Covariates_LF) <- colnames(pvals_Covariates_LF)
rownames(pvalsadj_Covariates_LF) <- rownames(pvals_Covariates_LF)


fdr <- 0.01
pvalDF <- melt(pvalsadj_Covariates_LF, varnames = c("covariate", "LF"), value.name = "p.adj")
pvalDF$LF <- as.factor(pvalDF$LF)
pvalDF$neg.log.p.adj <- -log(pvalDF$p.adj )
pvalDF$imp <- pvalDF$neg.log.p.adj > -log(fdr)
pvalDF$covariate <- ifelse(pvalDF$imp, as.character(pvalDF$covariate), "other")

cols <- c(RColorBrewer::brewer.pal(8,"Dark2"), RColorBrewer::brewer.pal(9,"Set1")) [1:length(unique(pvalDF$covariate))]
names(cols) <- unique(pvalDF$covariate)
cols["other"] <-"gray"

ggplot(pvalDF, aes(x= LF, y=neg.log.p.adj,col=covariate))+geom_point() + 
  geom_hline(yintercept = -log(fdr), linetype="dotted", col="red") + 
    ggrepel::geom_text_repel(data = filter(pvalDF, imp), aes(label = covariate, col = covariate), segment.alpha=0.2,
                             min.segment.length = unit(0.5, "lines"), nudge_x=0.5,  nudge_y = 20, show.legend=F) +
  coord_flip() + ylab("Neg. log p-value (adjusted)") + xlab("Latent Factor") +
  theme_bw() +geom_text(aes(y= -log(fdr), hjust=-0.1, x=ncol(Z)-1, label=paste("FDR", fdr*100,"%")), col="red") +
  scale_color_manual(values=cols) 
```

