---
title: "MOFA - Reviewer 1.3"
author: "Britta Velten"
date: "16/01/2018"
output:
  BiocStyle::html_document:
    toc: true
---

"In the analysis of the chronic lymphocytic leukemia data, not much information is provided about the choice of K in the three-mean clustering of samples based on factor 1 (Figure 3A). Was there a formal test about the goodness-of-fit of a three- versus e.g. two-mean clustering? The IZ and HZ clusters do not show a large difference in gene expression values (Figure 3C) or Drug response (Figure 3E). ""
 
```{r, message=F, warning=F}
library(MOFAtools)
library(dplyr)
library(reshape2)
library(pheatmap)
library(survival)
library(gridExtra)
library(ggplot2)
library(GGally)
library(magrittr)
library(Hmisc)
library(cowplot)
library(beeswarm)
# library(pace)
knitr::opts_chunk$set(dev=c('png', 'pdf'), fig.path = 'IGHV_whichK')
```

# Get imported model and related data
Prepared in CLL_import_technical.Rmd
```{r}
load("out_import_technical_flipped.RData")

#output path for plots
plotdir <- "~/Documents/MOFA/CLL_MOFA_data/Analysis/code4manuscript/IGHV_whichK"
if(!dir.exists(plotdir)) dir.create(plotdir)

# idx for IGHV latent factor
idx_ighv <- "1"

# helper function for some plots
source("plottingUtils.R")
```


```{r}
beeswarm::beeswarm(Z[,idx_ighv])
# cluster using kmeans to determine 2 IGHV factor/cluster
set.seed(32180)
ZMC <- kmeans(Z[,idx_ighv], 3, nstart=1000, iter.max = 1000)
ZMC <- ZMC$cluster
ZMC 
ZMC <- ifelse(ZMC==1, "HZ", ifelse(ZMC==2, "LZ", "IZ"))

```


# Test which K is optimal
```{r bic}
library(ClusterR)
Optimal_Clusters_KMeans(Z[,idx_ighv, drop=F], 5, criterion="BIC")
```

```{r}
# Optimal_Clusters_KMeans(Z[,idx_ighv, drop=F], 5, criterion="AIC")
# Optimal_Clusters_KMeans(Z[,idx_ighv, drop=F], 5, criterion="variance_explained")
# Optimal_Clusters_KMeans(Z[,idx_ighv, drop=F], 5, criterion="Adjusted_Rsquared")
# Optimal_Clusters_KMeans(Z[,idx_ighv, drop=F], 5, criterion="dissimilarity")
# Optimal_Clusters_KMeans(Z[,idx_ighv, drop=F], 5, criterion="WCSSE")
```


# Test difference in gene expression and drug response

## Gene expression
Comparison LZ to IZ: all 20 genes significant, comparison IZ to HZ 10/20 have p-value < 0.05.
```{r}
library(ggpubr)
weigths1 <- getWeights(model, "mRNA", idx_ighv)[[1]]
topGenes <- rownames(weigths1)[order(abs(weigths1), decreasing = T)][1:20]
cols <- c(LZ="navy", IZ="darkgoldenrod", HZ="darkred", "missing"="gray")

# test and plot
data <- melt(t(model@TrainData$mRNA[topGenes, ]), varnames = c("patient", "gene"))
data$cluster <- ZMC[data$patient]
pdf("~/Figure3supp_geneexpression.pdf", width=10, height=7)
gg <- ggplot(data, aes(x=cluster, y=value, group=cluster, fill=cluster)) +geom_boxplot()  +
    scale_fill_manual(values=cols) + ylab("Normalized expression")  + 
    stat_compare_means(comparisons=list(c("HZ","IZ"), c("IZ", "LZ")), method="t.test", label="p.signif", hide.ns=T) +
  facet_wrap(~gene, nrow =4, strip.position="top") + theme(panel.spacing = unit(2, "lines")) + ylim(c(0,20))
gg
dev.off()

# glist <- lapply(topGenes, function(gene){
#   df <- data.frame(value=model@TrainData$mRNA[gene,], cluster=ZMC)
#   gg <- ggplot(df, aes(x=cluster, y=value, group=cluster, fill=cluster)) +geom_boxplot()  +
#     scale_fill_manual(values=cols) + xlab("Normalized expression") +ggtitle(gene) + 
#     stat_compare_means(comparisons=list(c("HZ","IZ"), c("IZ", "LZ")), method="t.test")
# })
# Reducepvals_ge
# names(which(pvals_ge<0.05))
```

## Drug responses
```{r}
weigths1 <- getWeights(model, "Drugs", idx_ighv)[[1]]
# topDrugs <- rownames(weigths1)[order(abs(weigths1), decreasing = T)][1:20]
topDrugs <- rownames(weigths1)[grepl("dasatinib|AZD7762",rownames(weigths1))]

data <- model@TrainData$Drugs[topDrugs,]
data <- melt(data, varnames=c("drugconc","patient"))
data %<>% mutate(drugnm = sub("_[1-5]","",drugconc),
                 concid = substr(drugconc,nchar(as.character(drugconc)),nchar(as.character(drugconc))),
                 cluster=ZMC[patient])
data('conctab', package="pace")
data('drugs', package="pace")
data$drugid <- sapply(data$drugnm, function(dr) rownames(drugs[drugs$name==dr,, drop=F]))

data$conc <- sapply(1:nrow(data), function(r) {
  dd <- data[r,, drop=F]
  paste(conctab[dd$drugid,paste0("c",dd$concid)])
})

pdf("dasatinib_siginif.pdf")
ggplot(filter(data, drugnm=="dasatinib"), aes(x=cluster, group=cluster, y=value, fill=cluster)) + 
  geom_boxplot()+  scale_fill_manual(values=cols) +
  stat_compare_means(comparisons=list(c("HZ","IZ"), c("IZ", "LZ")), method="t.test") +facet_wrap(~conc)+
  ylim(c(0,1.5)) + ylab("Viability [%]") + ggtitle("dasatinib")
dev.off()

pdf("AZD7762_siginif.pdf")
ggplot(filter(data, drugnm=="AZD7762"), aes(x=cluster, group=cluster, y=value, fill=cluster)) + 
  geom_boxplot()+  scale_fill_manual(values=cols) +
  stat_compare_means(comparisons=list(c("HZ","IZ"), c("IZ", "LZ")), method="t.test") +facet_wrap(~conc)+
  ylim(c(0,1.5)) + ylab("Viability [%]") + ggtitle("AZD7762")
dev.off()

# test HZ vs IZ
pvals_dr <- sapply(topDrugs, function(drug){
  df <- data.frame(value=model@TrainData$Drugs[drug,], cluster=ZMC)
  gg <- ggplot(df, aes(x=cluster, y=value, group=cluster)) +geom_boxplot()
  # print(gg)
  ttout <- t.test(model@TrainData$Drugs[drug,ZMC=="HZ"],model@TrainData$Drugs[drug,ZMC=="IZ"])
  pval <- ttout$p.value
})
pvals_dr
sum(pvals_dr < 0.05)
hist(pvals_dr)
sum(p.adjust(pvals_dr, method="BH") < 0.1)
```

