---
title: "Compare imputation performance"
author: "Britta Velten"
date: '`r format(Sys.time(), "%d %B, %Y")`'
output:
  BiocStyle::html_document:
    toc: true
---

```{r, message=F, warning=F}
library(MOFAtools)
library(dplyr)
library(reshape2)
library(pheatmap)
library(data.table)
library(ggplot2)
```


# Source script performing imputations
```{r}
source("../Impute.R")
```


#In/Out options and define setting
The directories shoudl point to the directories where the GFA and MOFA fits are stored and to the directory of the (unmasked) data.
```{r}
out_mofa <- "/Users/bvelten/Documents/MOFA/GFA_imputation_models/CLL/imputation/atRandom/mofa/grid/"
out_gfa <- "/Users/bvelten/Documents/MOFA/GFA_imputation_models/CLL/imputation/atRandom/gfa/grid/"
dataDir <-"/Users/bvelten/Documents/MOFA/CLL_MOFA_data/views/minView=all/"
outdir <- "out/"
if(!dir.exists(outdir)) dir.create(outdir)
```

#Get MOFA models for imputation and masking parameters
```{r}
MOFAfiles <- list.files(out_mofa)
MOFAfiles <- MOFAfiles[grepl("0.05_", MOFAfiles)|grepl("0.5_", MOFAfiles)]
pattern <- paste0("_",1:20, "_",collapse ="|")
MOFAfiles <- MOFAfiles[grepl(pattern, MOFAfiles)]
length(MOFAfiles)
ListMOFA <- lapply(MOFAfiles[grepl(".hdf5",MOFAfiles)], function(run) loadModel(file.path(out_mofa, run)))

# check number of factors
K <- sapply(seq_along(ListMOFA), function(i) getDimensions(ListMOFA[[i]])$K)

# name models
names(ListMOFA) <- gsub(".hdf5", "",MOFAfiles[grepl(".hdf5",MOFAfiles)])

# collect experiment infos in dataframe
run<-sapply(strsplit(MOFAfiles, "_"), function(l) sub(".rds","",l[3]))
ELB <- sapply(ListMOFA, function(model) max(model@TrainStats$elbo, na.rm = T))
missing <- sapply(ListMOFA, function(m) sum(is.na(m@TrainData$viab))/prod(dim(m@TrainData$viab)))
labelmissing <- sapply(strsplit(names(missing), "_"), function(l) l[1])
labelK <- sapply(strsplit(names(missing), "_"), function(l) l[2])
stopifnot(all(K==labelK))
df <- data.frame(K=K,true=missing, name = names(missing), label=labelmissing, run=run)
rownames(df) <-NULL
head(df)
```

#Get GFA models for imputation
```{r}
GFAfiles <- list.files(out_gfa)
GFAfiles <- GFAfiles[grepl("0.05_", GFAfiles)|grepl("0.5_", GFAfiles)]
pattern <- paste0("_K",1:20, "_",collapse ="|")
GFAfiles <- GFAfiles[grepl(pattern, GFAfiles)]
length(GFAfiles)

ListGFA <- lapply(GFAfiles[grepl(".rds",GFAfiles)], function(run) readRDS(file.path(out_gfa, run)))
names(ListGFA) <- gsub(".rds|K", "",GFAfiles[grepl(".rds",GFAfiles)])

# check number of factors
K <- sapply(ListGFA, function(m) ncol(m$X))

# only include settings where mofa and gfa were fit
commonModels <- intersect(names(ListMOFA), names(ListGFA) )
ListMOFA <- ListMOFA[commonModels]
ListGFA <- ListGFA[commonModels]
all(names(ListMOFA)== names(ListGFA))

# check the views have the same ordering: NO need to reorder
all(rownames(ListGFA[[1]]$W) == Reduce(c,featureNames(ListMOFA[[1]])))

idx_MOFA <- which(!Reduce(c,featureNames(ListMOFA[[1]])) %in% rownames(ListGFA[[1]]$W))
idx_GFA <- which(!rownames(ListGFA[[1]]$W) %in% Reduce(c,featureNames(ListMOFA[[1]])))

# different names (e.g. "." instead of "-")

if(length(idx_MOFA)>0 | length(idx_GFA)>0){
# change a name in MOFA
rownames(ListGFA[[1]]$W)[idx_GFA]
(Reduce(c,featureNames(ListMOFA[[1]])))[idx_MOFA]
for(i in seq_along(ListMOFA))
featureNames(ListMOFA[[i]])$mut[featureNames(ListMOFA[[i]])$mut=="RP11-766F14.2"] <- "RP11.766F14.2"
}

# MOFA to GFA names mapping (should be the same for all models , but double-check)
GFA_to_MOFA_feautres <- sapply(seq_along(ListMOFA), function(i) match(Reduce(c,featureNames(ListMOFA[[i]])), rownames(ListGFA[[i]]$W)))
all(apply(GFA_to_MOFA_feautres,1, function(r) length(unique(r))==1))                           

GFA_to_MOFA_feautres <- GFA_to_MOFA_feautres[,1]
any(is.na(GFA_to_MOFA_feautres))
```

#Get true data
```{r}
views <- names(ListMOFA[[1]]@TrainData)
allData <- paste(views, ".txt", sep="")
DataList <- lapply(allData, function(file) read.table(file.path(dataDir, file)))
names(DataList) <- views
rownames(DataList[[2]])==sampleNames(ListMOFA[[1]])
rownames(DataList[[2]])==rownames(ListGFA[[1]]$X)
DataList_scaled <- lapply(DataList, function(dd) scale(dd,center=TRUE, scale=FALSE))
```


#Impute

## Using all views
Impute and evaluate imputations
```{r}
  imp_results <- lapply(names(ListMOFA), function(modelnm){
  print(modelnm)
  mofa.tmp <- ListMOFA[[modelnm]]
  gfa.tmp <- ListGFA[[modelnm]]
  imp.out <- ImputeComparison(mofa.tmp, gfa.tmp, DataList_scaled, GFA_to_MOFA_feautres)
  imp.out$MSEdf$modelName <- modelnm
  imp.out
}) 
names(imp_results) <- names(ListMOFA)
benchmark_results<-lapply(names(imp_results), function(impnm) {
  dftmp <- imp_results[[impnm]]$MSEdf
  dftmp$param_PercNA <- filter(df, name==impnm)$label
  dftmp$run <- filter(df, name==impnm)$run
  dftmp$K <- filter(df, name==impnm)$K
  dftmp
}
  ) %>% bind_rows()
benchmark_results$ELB <- ELB[benchmark_results$modelName]

# save imputation results
save(benchmark_results, file=file.path(outdir,"imputation_random_drug_grid.RData"))
```

# SessionInfo
```{r}
sessionInfo()
```

