---
title: "Imputation for data with complete missingness in one view"
author: "Britta Velten"
date: "10/08/2017"
output:
  BiocStyle::html_document:
    toc: true
---

```{r, message=F, warning=F}
setwd("/Users/bvelten/Documents/MOFA/CLL_MOFA_data/Analysis/code4manuscript/")
devtools::load_all("/Users/bvelten/Documents/MOFA/MOFApackage/scGFA/MOFAtools/")
library(dplyr)
library(reshape2)
library(pheatmap)
library(data.table)
library(ggplot2)
source("~/Documents/MOFA/CLL_MOFA_data/Analysis/imputation/ImputeBenchmark.R")
```

# Overview of the model

For imputation, we include n=121 samples into the model, which have complete information in the following views

- RNAseq: mRNA (after removal of genes with low counts, variance stabilized, top 5000 most variable)
- methylaltion (transformed to M values and filtered to top 1% most variable CpG sites)
- viabilities (all drugs and concentrations)

Additional, mutations are included as forth view despite missing values.

Pre-processing scripts from the orginial data files are found in the data folder/views.
Some values are masked to be imputed here. The fraction of masked entries can be specified as parameter for the python run_FullCases script.

#In/Out options 
```{r}
# outDir<-"/Users/bvelten/Documents/MOFA/CLL_MOFA_data/out/imputation/FullCases/10Aug/"
outDir<-"/Users/bvelten/Documents/MOFA/CLL_MOFA_data/out/imputation/FullCases/23Aug/"
dataDir<-"/Users/bvelten/Documents/MOFA/CLL_MOFA_data/views/minView=all/"
ktype <- "learnK" # k5
```

#Get all models for imputation
```{r}
allRuns <- list.files(outDir)

#first look at learnk
# allRuns <- allRuns[grepl(ktype,allRuns)]
ListModels <- lapply(allRuns[grepl(".hdf5",allRuns)], function(run) loadModel(file.path(outDir, run), sortFactors = F))
sapply(seq_along(ListModels), function(i) getDimensions(ListModels[[i]])$K)

# remove values for samples missing a view if the factor is unique to this view
ListModels <- lapply(ListModels, detectPassengers)
sapply(lapply(seq_along(ListModels), function(i) ListModels[[i]]@Expectations$Z$E), function(z) any(is.na(z)))

# name models
names(ListModels) <- allRuns[grepl(".hdf5",allRuns)]

# get info on each model
ELB <- sapply(ListModels, function(model) max(model@TrainStats$elbo, na.rm = T))
missing <- sapply(ListModels, function(m) unique(apply(m@TrainData$viab,1, function(c) sum(is.na(c)))))
labelmissing <- sapply(strsplit(names(missing), "_"), function(l) l[1])
df <- data.frame(true=missing, name = names(missing), label=labelmissing)
rownames(df) <-NULL
df
```

#Get true data
```{r}
views <- names(ListModels[[1]]@TrainData)
allData <- paste(views, ".txt", sep="")
DataList <- lapply(allData, function(file) read.table(file.path(dataDir, file))) #fread does not support rownames?
names(DataList) <- views
```

#Impute
Note: The imputeR package has a lot of bugs, better re-implement. E.g. if not all feautres have a missing value breaks (is the case for 5%).
kNN <-> different scales?

Impute
```{r}
imp_results<-lapply(names(ListModels), function(modelnm){
  modeltmp <- ListModels[[modelnm]]
  imp.out <- ImputeBenchmark(modeltmp,DataList)
  imp.out$MSEdf$modelName <- modelnm
  imp.out
}) 

benchmark_results<-lapply(1:length(imp_results), function(impidx) {
  dftmp<-imp_results[[impidx]]$MSEdf
}
  ) %>% bind_rows()

benchmark_results$ELB <- ELB[benchmark_results$modelName]
# save(benchmark_results, file=paste("benchmark_results_10Aug_full_viab_rmPassengers_",ktype,".RData", sep=""))
save(benchmark_results, file=paste("benchmark_results_23Aug_full_viab_rmPassengers_",ktype,".RData", sep=""))
```

