---
title: "GFA_Imputation"
date: "14 August 2017"
output:
  BiocStyle::html_document:
    toc: true
---

```{r, message=F, warning=F}
devtools::load_all("/Users/ricard/mofa/MOFAtools")
# library(MOFAtools)
library(dplyr)
library(reshape2)
library(pheatmap)
library(data.table)
library(ggplot2)
```

# Overview of the model

For imputation, we include n=121 samples into the model, which have complete information in the following views

- RNAseq: mRNA (after removal of genes with low counts, variance stabilized, top 5000 most variable)
- methylaltion (transformed to M values and filtered to top 1% most variable CpG sites)
- viabilities (all drugs and concentrations)

Additional, mutations are included as forth view despite missing values.

Pre-processing scripts from the orginial data files are found in the data folder/views.
Some values are masked to be imputed here. The fraction of masked entries can be specified as parameter for the python run_AtRandom script.

# Source script performing imputations
```{r}
source("ImputeBenchmark.R")
```


#In/Out options and define setting (ktype/view)
```{r}
ktype <- ""
view <- "drug"
run_date <- "6Sep"
```

<!-- ```{r} -->
<!-- if(view=="expr"){ -->
<!-- outDir <- paste0("/Users/bvelten/Documents/MOFA/CLL_MOFA_data/out/imputation/missingAtRandom/", run_date) -->
<!-- } else if(view=="viab") { -->
<!--   outDir <- paste0("/Users/bvelten/Documents/MOFA/CLL_MOFA_data/out/imputation/missingAtRandom/",run_date) -->
<!-- } else stop("No masked models for given view") -->

<!-- dataDir <-"/Users/bvelten/Documents/MOFA/CLL_MOFA_data/views/minView=all/" -->
<!-- ``` -->

```{r}
outDir <- paste0("/Users/ricard/data/CLL/out/imputation/missingAtRandom/",run_date)
dataDir <-"/Users/ricard/data/CLL/views/minView=all"
```

#Get all models for imputation
```{r}
allRuns <- list.files(outDir)
ListModels <- lapply(allRuns[grepl(".hdf5$",allRuns)], function(run) loadModel(file.path(outDir, run), sortFactors = F))

# check number of factors
# sapply(seq_along(ListModels), function(i) getDimensions(ListModels[[i]])$K)

# remove values for samples missing a view if the factor is unique to this view (now taken care of in loadModel)
# ListModels <- lapply(ListModels, detectPassengers)
# sapply(lapply(seq_along(ListModels), function(i) ListModels[[i]]@Expectations$Z$E), function(z) any(is.na(z)))

# name models
names(ListModels) <- allRuns[grepl(".hdf5$",allRuns)]

# collect experiment infos in dataframe
run<-sapply(strsplit(allRuns, "_"), function(l) sub(".hdf5","",l[7]))
ELB <- sapply(ListModels, function(model) max(model@TrainStats$elbo, na.rm = T))
missing <- sapply(ListModels, function(m) sum(is.na(m@TrainData$viab))/prod(dim(m@TrainData$viab)))
labelmissing <- sapply(strsplit(names(missing), "_"), function(l) l[1])
df <- data.frame(true=missing, name = names(missing), label=labelmissing)
rownames(df) <-NULL
# df
```

#Get true data
```{r}
views <- names(ListModels[[1]]@TrainData)
allData <- paste(views, ".txt", sep="")
DataList <- lapply(allData, function(file) read.table(file.path(dataDir, file)))
names(DataList) <- views
```


#Impute

## Using all views
Impute and evaluate imputations
```{r}
imp_results<-lapply(names(ListModels), function(modelnm){
  modeltmp <- ListModels[[modelnm]]
  imp.out<-ImputeBenchmark(modeltmp,DataList)
  imp.out$MSEdf$modelName <- modelnm
  imp.out
}) 

benchmark_results<-lapply(1:length(imp_results), function(impidx) {
  dftmp<-imp_results[[impidx]]$MSEdf
  dftmp$param_PercNA <- labelmissing[impidx]
  dftmp$run <- run[impidx]
  dftmp
}) %>% bind_rows()
benchmark_results$ELB <- ELB[benchmark_results$modelName]

# save imputation results
save(benchmark_results, file=paste("benchmark_results_",run_date,"_random_",view,"_rmPassengers_",ktype,".RData", sep=""))
```

```{r, eval=F, echo=F}
# Impute using only masked view:

# imp_results<-lapply(names(ListModels), function(modelnm){
#   modeltmp <- ListModels[[modelnm]]
#   imp.out<-ImputeBenchmarkOneView(modeltmp,"viab",DataList)
#   imp.out$MSEdf$modelName <- modelnm
#   imp.out
# }) 
# 
# benchmark_results<-lapply(1:length(imp_results), function(impidx) {
#   dftmp<-imp_results[[impidx]]$MSEdf
#   dftmp$param_PercNA <- labelmissing[impidx]
#   dftmp$run <- run[impidx]
#   dftmp
# }
#   ) %>% bind_rows()
# 
# benchmark_results$ELB <- ELB[benchmark_results$modelName]
# save(benchmark_results, file=paste("SingleViewBenchmark_results_10Aug_random_viab_rmPassengers_",ktype,".RData", sep=""))
```

