---
title: "MOFA - Results using one view only or excluding one view"
author: "Britta Velten"
date: '`r format(Sys.time(), "%d %B, %Y")`'
output:
  BiocStyle::html_document:
    toc: true
---

```{r, message=F, warning=F}
library(MOFAtools)
library(dplyr)
library(reshape2)
library(pheatmap)
library(survival)
library(gridExtra)
library(ggplot2)
library(GGally)
library(magrittr)
library(Hmisc)
library(glmnet)
library(cowplot)
knitr::opts_chunk$set(fig.path = "figs_maskingAnalysis/", dev=c('png','pdf'))
```

```{r}
io <- list()
io$outdir_one <- "out/oneview"
io$outdir_three <- "out/threeview"
```

# Load orignial model
```{r}
load("../out_import_technical_flipped.RData")
completeCases <- Reduce(intersect,lapply(data, function(dd) colnames(dd)[apply(dd,2,function(p) !all(is.na(p)))]))
```

# Get survival data
```{r}
survivalData <- as.matrix(patmeta[,c("T5","treatedAfter")])
survivalData<- survivalData[!is.na(survivalData[,1]),]
colnames(survivalData)<-c("time", "status")
```

# Colors
Define colors for barplot
```{r}
# define colors
cols4survival_oneview <- cols4survival_threeview <- c("#E69F00","#D55E00","#009E73", "#56B4E9",  "#999999")
names(cols4survival_oneview) <- c("Methylation", "Mutations","mRNA",  "Drugs", "joint")
names(cols4survival_threeview) <- c(paste0("no_",c("Methylation", "Mutations","mRNA",  "Drugs")), "joint")
```

# One view models
## Load one view models
```{r}
files <- list.files(io$outdir_one)
OneViewModels <- lapply(files, function(fnm){
  m <- loadModel(file.path(io$outdir_one,fnm))
  MOFAtools:::.updateOldModel(m)
} )
names(OneViewModels) <- sub(".hdf5", "", files)
```

## Get factors
```{r}
AllFactors_oneView <- lapply(seq_along(OneViewModels), function(m){
  Zm <- getFactors(OneViewModels[[m]], include_intercept = FALSE)
  colnames(Zm) <- paste(names(OneViewModels)[m], colnames(Zm), sep="_")
  Zm[Zm==0] <- NA
  Zm
}) 
names(AllFactors_oneView) <- names(OneViewModels)
AllFactors_oneView_mat <- Reduce(cbind,AllFactors_oneView)
```


## Correlation to original factors
For each original factor look at maximal correlation to new factors from the subsetted model.
```{r cor_1view}
AllFactors_oneView_sub <- lapply(AllFactors_oneView, function(Zm) Zm[completeCases,])
max_cor <- sapply(rev(AllFactors_oneView_sub), function(Zm) apply(abs(cor(Z[completeCases,-1], Zm, use="complete.obs")),1,max))
df_cor_1view <- melt(max_cor, varnames = c("Factor", "Omic"), value.name = "corr")
gg_corr_oneview <- ggplot(df_cor_1view, aes(x=factor(Factor, levels = 1:10),
                              y=factor(Omic, levels = rev(c("Mutations", "mRNA", "Methylation", "Drugs"))),
                              fill=corr)) + 
  geom_tile() + 
  guides(fill=guide_colorbar(title="Absolute \n Correlation")) +
  # scale_fill_gradient(low="white", high="red") +
    scale_fill_gradientn(colors=c("white","orange","red")) +
  xlab("Original factor") + ylab("Data for training")
```


## Survival prediction
Harrals C-Index is used as performance measure, 5-fold stratified cross-validation. Only on samples having all omics.
```{r, echo=FALSE}
getCIndex <- function(AllFactors){
  
  # subset to common patients, measured in all assays
  # use only those 121 patients who have all the data to get information on which dataset is most informative not confounded by number of available patients
  commonPats <- intersect(rownames(Z), rownames(survivalData))
  commonPats <- intersect(commonPats, completeCases)
  print(length(commonPats))

  # add clinnical covariates as predictors
  Zcommon <- Z[commonPats,-1]
  AllFactors_common <- lapply(AllFactors, function(Zm) Zm[commonPats,])
  survivalData_common <- survivalData[commonPats,]
  print(paste("uncensored:",sum(survivalData_common[,2])))

  # construct survival object
  SurvObject <-Surv(survivalData_common[,1],survivalData_common[,2])
  
  #stratified CV to include same proportion of uncensored cases
  set.seed(1290)
  uncensored <- SurvObject[,"status"]==1
  cv_ix <- dismo::kfold(1:length(commonPats), k=5, by=uncensored)
  
  # add covariates from forest plot
  FeatureList <- AllFactors_common
  FeatureList$Mutations <-  FeatureList$Mutations[,1, drop=FALSE] # only first four factors due to singularity issue
  FeatureList$joint = Zcommon
  
  # fit a cox model and predict on left-out fold
  fit <- lapply(unique(cv_ix), function(i){
    
      #fit coxph for reduced
      c = lapply(FeatureList, function(x) coxph(SurvObject[cv_ix!=i,]  ~ as.matrix(x)[cv_ix!=i,]))
      p = mapply(function(x,y) as.matrix(x[cv_ix==i,]) %*% coef(y), FeatureList, c)
      
      CI=apply(-p,2, rcorr.cens, SurvObject[cv_ix==i])[1,]
      list(CI=CI,c=c)
  })

    # get cross-validated CI
    concordanceCV <- sapply(fit, function(l) l$CI)
    colnames(concordanceCV) <- unique(cv_ix)
    df_hc <- melt(concordanceCV, varnames = c("predictor", "cv_idx"), value.name = "CI")
    df_hc$survType <- "Time to next treatment"

# calculate summary statistics across folds
summaryHc <- aggregate(df_hc$CI,
    by = list(predictors = df_hc$predictor, survType=df_hc$survType), 
    FUN = function(x) c(mean = mean(x), sd = sd(x),
                        n = length(x)))
summaryHc <- do.call(data.frame, summaryHc)
summaryHc$se <- summaryHc$x.sd / sqrt(summaryHc$x.n)

summaryHc
}
```

```{r}
summaryHc_oneview <- getCIndex(AllFactors_oneView)
```

Make barplot
```{r barplot_1view}
limits <- aes(ymax = summaryHc_oneview$x.mean + summaryHc_oneview$se,
              ymin = summaryHc_oneview$x.mean - summaryHc_oneview$se)

# barplot
gg_bar_oneview <- ggplot(summaryHc_oneview, aes(x=predictors, y=x.mean, fill=predictors, group=predictors))+
    geom_bar(stat = "identity") + 
   coord_cartesian(ylim = c(0.5,0.85)) +
  scale_fill_manual(values=cols4survival_oneview) + 
  geom_errorbar(limits, position = "dodge", width = 0.25)+
  theme(axis.text.x = element_text(angle = 60, vjust=1,hjust = 1,  lineheight=-2), 
        strip.background = element_rect(fill="white"),
        strip.text = element_blank(),
        legend.position= "none",
        plot.title = element_text(size=18)) +
  ylab("Harrell's C") + xlab("")
```


# Three view models
## Load three view models
```{r}
files <- list.files(io$outdir_three)
ThreeViewModels <- lapply(files, function(fnm){
  m <- loadModel(file.path(io$outdir_three,fnm))
  MOFAtools:::.updateOldModel(m)
} )
names(ThreeViewModels) <- sub(".hdf5", "", files)
```

## Get factors
```{r}
AllFactors_threeView <- lapply(seq_along(ThreeViewModels), function(m){
  Zm <- getFactors(ThreeViewModels[[m]], include_intercept = FALSE)
  colnames(Zm) <- paste(names(ThreeViewModels)[m], colnames(Zm), sep="_")
  Zm[Zm==0] <- NA
  Zm
}) 
names(AllFactors_threeView) <- names(ThreeViewModels)
AllFactors_threeView_mat <- Reduce(cbind,AllFactors_threeView)
```


## Correlation to original factors
For each original factor look at maximal correlation to new factors from the subsetted model.
```{r cor_3view}
AllFactors_threeView_sub <- lapply(AllFactors_threeView, function(Zm) Zm[completeCases,])
max_cor <- sapply(rev(AllFactors_threeView_sub), function(Zm) apply(abs(cor(Z[completeCases,-1], Zm, use="complete.obs")),1,max))
df_cor_3view <- melt(max_cor, varnames = c("Factor", "Omic"), value.name = "corr")
gg_corr_threeview <- ggplot(df_cor_3view, aes(x=factor(Factor, levels = 1:10),
                                        y=factor(Omic, levels = rev(paste0("no_",c("Mutations", "mRNA", "Methylation", "Drugs")))),
                                        fill=corr)) + 
  geom_tile() + 
  guides(fill=guide_colorbar(title="Absolute \n Correlation")) +
  # scale_fill_gradient(low="white", high="red") +
    scale_fill_gradientn(colors=c("white","orange","red")) +
  xlab("Original factor") + ylab("Data for training")
```


## Survival prediction
```{r}
summaryHc_threeView <- getCIndex(AllFactors_threeView)
```

Make barplot
```{r barplot_3view}
limits <- aes(ymax = summaryHc_threeView$x.mean + summaryHc_threeView$se,
              ymin = summaryHc_threeView$x.mean - summaryHc_threeView$se)

# barplot
gg_bar_threeview <- ggplot(summaryHc_threeView, aes(x=predictors, y=x.mean, fill=predictors, group=predictors))+
    geom_bar(stat = "identity") + 
   coord_cartesian(ylim = c(0.5,0.85)) +
  scale_fill_manual(values=cols4survival_threeview) + 
  geom_errorbar(limits, position = "dodge", width = 0.25)+
  theme(axis.text.x = element_text(angle = 60, vjust=1,hjust = 1,  lineheight=-2), 
        strip.background = element_rect(fill="white"),
        strip.text = element_blank(),
        legend.position= "none",
        plot.title = element_text(size=18)) +
  ylab("Harrell's C") + xlab("")
```


# Arrange figures
```{r fig_SXX_subviews, fig.height=7, fig.width=10}
gg_c1 <- plot_grid(gg_corr_oneview, gg_corr_threeview,ncol=1, labels=c("a","c"), align="hv", axis = "lb", label_size = 22)
gg_c2 <- plot_grid(gg_bar_oneview, gg_bar_threeview, ncol=1, labels=c("b", "d"), align="hv", axis = "lb", label_size = 22)
plot_grid(gg_c1, gg_c2, nrow=1)
# plot_grid(gg_corr_oneview, gg_bar_oneview, gg_corr_threeview, gg_bar_threeview,nrow=2, labels=c("a","b", "c", "d"), align="hv", axis = "lb")
```

